---
title: "eCLIPSE: Building the betAS ASfile"
subtitle: "Differential splicing analysis with betAS, from a VAST-TOOLS inclusion table to the Event.ID / dPSI table eCLIPSE_full() and eCLIPSE_full_IR() expect"
author:
  - name: "Alexandre Afonso^[Disease Transcriptomics Group, Instituto de Medicina Molecular, Faculdade de Medicina da Universidade de Lisboa]"
    email: alexandreafonso@medicina.ulisboa.pt
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  rmdformats::downcute:
    downcute_theme: "chaos"
---

# Introduction

This notebook takes a VAST-TOOLS inclusion table and produces the `ASfile` that `eCLIPSE_full()`/`eCLIPSE_full_IR()` expect: a table with an `Event.ID` column and a `dPSI` column, computed with the [betAS](https://github.com/DiseaseTranscriptomicsLab/betAS) package. The example below compares a control group against an shRNA-knockdown group, a typical ENCODE-style eCLIP experimental design.

```{r}
library(betAS)
library(data.table)
```

# Loading and filtering the dataset

```{r}
dataset <- getDataset(pathTables = "~/path/to/your/project/INCLUSION_LEVELS_FULL-hg38-25-v251.tab.gz", tool = "vast-tools")
dataset <- getEvents(dataset, tool = "vast-tools")

# Keep exon-skipping (S) and intron-retention (IR) events
dataset_filtered <- filterEvents(dataset, types = c("S", "IR"), N = 0)
dataset_filtered$PSI  <- na.omit(dataset_filtered$PSI)
dataset_filtered$Qual <- dataset_filtered$Qual[dataset_filtered$Qual$EVENT %in% dataset_filtered$PSI$EVENT, ]

# Two groups: control and shRNA-knockdown samples
samples_control <- c("control_rep1", "control_rep2")
samples_shRNA   <- c("shRNA_rep1", "shRNA_rep2")

dataset_control <- dataset_filtered
dataset_control[["Samples"]] <- samples_control
dataset_control[["PSI"]]  <- dataset_control$PSI[,  c(1:6, which(colnames(dataset_control$PSI)  %in% samples_control))]
dataset_control[["Qual"]] <- dataset_control$Qual[, c(1:6, which(colnames(dataset_control$Qual) %in% paste0(samples_control, ".Q")))]

dataset_shRNA <- dataset_filtered
dataset_shRNA[["Samples"]] <- samples_shRNA
dataset_shRNA[["PSI"]]  <- dataset_shRNA$PSI[,  c(1:6, which(colnames(dataset_shRNA$PSI)  %in% samples_shRNA))]
dataset_shRNA[["Qual"]] <- dataset_shRNA$Qual[, c(1:6, which(colnames(dataset_shRNA$Qual) %in% paste0(samples_shRNA, ".Q")))]
```

# Filtering to informative events

There are two ways to filter `dataset_control`/`dataset_shRNA` down to the events worth testing. Pick **one** — both options below produce a same-named `_filtered` object, so running both just means the second one wins.

## Option A: betAS's own `alternativeEvents()`

Filters events to a PSI range, dropping events with near-0% or near-100% inclusion across all samples (which carry little information about a splicing change):

```{r}
dataset_control_filtered <- alternativeEvents(dataset_control, minPsi = 1, maxPsi = 99)
dataset_shRNA_filtered   <- alternativeEvents(dataset_shRNA,   minPsi = 1, maxPsi = 99)
```

## Option B: `NewFilterFunction()`

Not part of betAS itself. Instead of a PSI range, it keeps only events that appear in a separately-precomputed event list, `FullCassetteEvents.txt`. For information about its processing, please consult eCLIPSE-Supplementary.Rmd.

```{r}
NewFilterFunction <- function (filteredList, eventlist) 
{
  alternative <- list()
  psiTable <- filteredList$PSI
  qualTable <- filteredList$Qual
  originalColN <- ncol(psiTable)
  psi <- psiTable[, -c(1:6)]
  psiTable <- psiTable[psiTable$EVENT %in% eventlist$EVENT, ]
  qualTable <- qualTable[match(psiTable$EVENT, qualTable$EVENT), 
    ]
  psiTable <- psiTable[, c(1:originalColN)]
  qualTable <- qualTable[, c(1:originalColN)]
  alternative[[1]] <- psiTable
  alternative[[2]] <- qualTable
  alternative[[3]] <- table(psiTable$COMPLEX)
  alternative[[4]] <- colnames(psiTable)[-c(1:6)]
  names(alternative) <- c("PSI", "Qual", "EventsPerType", 
    "Samples")
  return(alternative)
}

```

```{r}
FullTrial <- fread("~/path/to/your/project/FullCassetteEvents.txt")

dataset_control_filtered <- NewFilterFunction(dataset_control, FullTrial)
dataset_shRNA_filtered   <- NewFilterFunction(dataset_shRNA,   FullTrial)
```

# Computing dPSI and building the ASfile

```{r}
cols_control <- convertCols(dataset_control_filtered$PSI, samples_control)
cols_shRNA   <- convertCols(dataset_shRNA_filtered$PSI,   samples_shRNA)

volcanoTable <- prepareTableVolcano(
  psitable       = dataset_control_filtered$PSI,   # control/shRNA filtered tables share the same events by construction
  qualtable      = dataset_control_filtered$Qual,
  npoints        = 500,
  colsA          = cols_control,
  colsB          = cols_shRNA,
  labA           = "control",
  labB           = "shRNA",
  basalColor     = "#89C0AE",
  interestColor  = "#E69A9C",
  maxDevTable    = maxDevSimulationN100,   # ships with the betAS package itself
  seed           = TRUE,
  CoverageWeight = FALSE
)

(volcano_plot <- plotVolcano(betasTable = volcanoTable,
                              labA = "control", labB = "shRNA",
                              basalColor = "#89C0AE", interestColor = "#E69A9C"))

ASfile <- volcanoTable[, c("EVENT", "GENE", "deltapsi", "Pdiff")]
colnames(ASfile) <- c("Event.ID", "Gene", "dPSI", "Pdiff")

write.table(ASfile, "~/path/to/your/project/my_betAS_table.txt", row.names = FALSE, col.names = TRUE)
```


