---
title: "eCLIPSE: RNA Binding Map Functions"
subtitle: "Second half of the RNA Binding Maps stage: eCLIPSE_full()/eCLIPSE_full_IR(), per-RBP binding heatmaps, and worked running examples"
author:
  - name: "Alexandre Afonso^[Disease Transcriptomics Group, Instituto de Medicina Molecular, Faculdade de Medicina da Universidade de Lisboa]"
    email: alexandreafonso@medicina.ulisboa.pt
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  rmdformats::downcute:
    downcute_theme: "chaos"
---

# Introduction

This is the second half of the RNA Binding Maps stage. The first half, reshaping `eCLIPSE_Exon.py`/`eCLIPSE_Intron.py`'s raw overlap output into the wide `rnamapfile` position matrix these functions expect, lives in `Eclip_position_matrix.Rmd`. Please run that first (see Section 7 of the tutorial). 

`eCLIPSE_full()` is the exon-skipping RNA-binding map function, and works for both the VastDB/VAST-TOOLS path and the rMATS path. `eCLIPSE_full_IR()` is the intron-retention RNA-binding map function.

```{r}
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(ggpubr)
  library(scales)
  library(reshape2)
})
```

# `eCLIPSE_full()`

```{r}
eCLIPSE_full <- function(rnamapfile, pvalthreshold=0.05, PSIthreshold=0.05, ASfile, rnaBP, metric = "FDR", plot = T, title = NULL, ylim_metric = NULL){

  # --- FIX: was ASfile <- ASfile[grepl("HsaEX", Event.ID)]                 ---
  # --- (VastDB-only; silently dropped every row for rMATS numeric IDs).    ---
  # --- Now format-agnostic: keep only events that actually have a row in  ---
  # --- the binding-metagenome table (rnamapfile), whatever ID scheme it   ---
  # --- uses (VastDB "HsaEX..." or rMATS integer IDs).                     ---
  ASfile <- ASfile[ASfile$Event.ID %in% rnamapfile$EVENTS, ]
  AS_events <- ASfile$Event.ID

  ASfile_filt_increased <- subset(ASfile, dPSI>PSIthreshold)
  ASfile_filt_decreased <- subset(ASfile, dPSI< -PSIthreshold)
  ASfile_filt_maintained <- subset(ASfile, !(dPSI > PSIthreshold) & !(dPSI < -PSIthreshold))

  #And now the eCLIPSE file
  rnamapfile_rbp_increased <- subset(rnamapfile, EVENTS %in% ASfile_filt_increased$Event.ID)
  rnamapfile_rbp_maintained <- subset(rnamapfile, EVENTS %in% ASfile_filt_maintained$Event.ID)
  rnamapfile_rbp_decreased <- subset(rnamapfile, EVENTS %in% ASfile_filt_decreased$Event.ID)

  rnamapfile_rbp_increased_rbp <- subset(rnamapfile_rbp_increased, RBP==rnaBP)
  rnamapfile_rbp_increased_nrbp <- subset(rnamapfile_rbp_increased, RBP!=rnaBP)
  rnamapfile_rbp_increased_nrbp <- rnamapfile_rbp_increased_nrbp[!duplicated(rnamapfile_rbp_increased_nrbp$EVENTS), ]
  rnamapfile_rbp_increased_nrbp[rnamapfile_rbp_increased_nrbp == 1] <- 0
  rnamapfile_rbp_maintained_rbp <- subset(rnamapfile_rbp_maintained, RBP==rnaBP)
  rnamapfile_rbp_maintained_nrbp <-subset(rnamapfile_rbp_maintained, RBP!=rnaBP)
  rnamapfile_rbp_maintained_nrbp <- rnamapfile_rbp_maintained_nrbp[!duplicated(rnamapfile_rbp_maintained_nrbp$EVENTS), ]
  rnamapfile_rbp_maintained_nrbp[rnamapfile_rbp_maintained_nrbp == 1] <- 0
  rnamapfile_rbp_decreased_rbp <- subset(rnamapfile_rbp_decreased, RBP==rnaBP)
  rnamapfile_rbp_decreased_nrbp <- subset(rnamapfile_rbp_decreased, RBP!=rnaBP)
  rnamapfile_rbp_decreased_nrbp <- rnamapfile_rbp_decreased_nrbp[!duplicated(rnamapfile_rbp_decreased_nrbp$EVENTS), ]
  rnamapfile_rbp_decreased_nrbp[rnamapfile_rbp_decreased_nrbp == 1] <- 0

  rnamapfile_rbp_increased <- rbind(rnamapfile_rbp_increased_rbp, rnamapfile_rbp_increased_nrbp)
  rnamapfile_rbp_decreased <- rbind(rnamapfile_rbp_decreased_rbp, rnamapfile_rbp_decreased_nrbp)
  rnamapfile_rbp_maintained <- rbind(rnamapfile_rbp_maintained_rbp, rnamapfile_rbp_maintained_nrbp)

  if (nrow(rnamapfile_rbp_increased) < 10 ||
      nrow(rnamapfile_rbp_maintained) < 10 ||
      nrow(rnamapfile_rbp_decreased) < 10) {
    print(paste0(rnaBP," does not have enough events."))
    return()
  }

  rnamapfile_rbp_increased <- rnamapfile_rbp_increased[,-c(1001,1002)]
  rnamapfile_rbp_increased <- apply(rnamapfile_rbp_increased, c(1, 2), as.numeric)
  rnamapfile_rbp_increased_comp <- colSums(rnamapfile_rbp_increased, na.rm = TRUE)
  rnamapfile_rbp_increased_incomp <- colSums(!is.na(rnamapfile_rbp_increased))

  rnamapfile_rbp_decreased <- rnamapfile_rbp_decreased[,-c(1001,1002)]
  rnamapfile_rbp_decreased <- apply(rnamapfile_rbp_decreased, c(1, 2), as.numeric)
  rnamapfile_rbp_decreased_comp <- colSums(rnamapfile_rbp_decreased, na.rm = TRUE)
  rnamapfile_rbp_decreased_incomp <- colSums(!is.na(rnamapfile_rbp_decreased))

  rnamapfile_rbp_maintained <- rnamapfile_rbp_maintained[,-c(1001,1002)]
  rnamapfile_rbp_maintained <- apply(rnamapfile_rbp_maintained, c(1, 2), as.numeric)
  rnamapfile_rbp_maintained_comp <- colSums(rnamapfile_rbp_maintained, na.rm = TRUE)
  rnamapfile_rbp_maintained_incomp <- colSums(!is.na(rnamapfile_rbp_maintained))

  #PLOTTING
  dfforvis <- data.frame(
    Name = as.numeric(names(rnamapfile_rbp_maintained_comp)),
    IncreasedEvents = as.numeric(rnamapfile_rbp_increased_comp+1),
    DecreasedEvents = as.numeric(rnamapfile_rbp_decreased_comp+1),
    MaintainedEvents = as.numeric(rnamapfile_rbp_maintained_comp+1)
  )

  pval_inc <- c()
  pval_dec <- c()
  oddsratio_inc <- c()
  oddsratio_dec <- c()
  for (i in 1:1000){
    inc_fish <- dfforvis[i,2]
    inc_fish_rev <- (rnamapfile_rbp_increased_incomp[i] - dfforvis[i,2])
    inc_fish_vs <- (dfforvis[i,4])
    inc_fish_vs_rev <- (rnamapfile_rbp_maintained_incomp[i])-(dfforvis[i,4])

    dec_fish <- dfforvis[i,3]
    dec_fish_rev <- (rnamapfile_rbp_decreased_incomp[i] - dfforvis[i,3])
    dec_fish_vs <- (dfforvis[i,4])
    dec_fish_vs_rev <- (rnamapfile_rbp_maintained_incomp[i])-(dfforvis[i,4])

    inc_df <- matrix(c(inc_fish, inc_fish_rev, inc_fish_vs, inc_fish_vs_rev), nrow = 2)
    inc_pval <- chisq.test(inc_df)$p.value
    inc_odds <- chisq.test(inc_df)$statistic

    dec_df <- matrix(c(dec_fish, dec_fish_rev, dec_fish_vs, dec_fish_vs_rev), nrow = 2)
    dec_pval <- chisq.test(dec_df)$p.value
    dec_odds <- chisq.test(dec_df)$statistic

    inc_pval <- -log10(p.adjust(inc_pval, method = "BH", n = 1000))
    dec_pval <- -log10(p.adjust(dec_pval, method = "BH", n = 1000))

    if ((inc_fish/rnamapfile_rbp_increased_incomp[i]) <=(inc_fish_vs/(rnamapfile_rbp_maintained_incomp[i]))){
      inc_pval <- inc_pval*-1
      inc_odds <- inc_odds*-1
    }
    if ((dec_fish/rnamapfile_rbp_decreased_incomp[i]) <=(dec_fish_vs/(+rnamapfile_rbp_maintained_incomp[i]))){
      dec_pval <- dec_pval*-1
      dec_odds <- dec_odds*-1
    }
    pval_inc <- c(pval_inc,inc_pval)
    oddsratio_inc <- c(oddsratio_inc, inc_odds)
    pval_dec <- c(pval_dec,dec_pval)
    oddsratio_dec <- c(oddsratio_dec, dec_odds)
  }

  dfforvis_pval <- data.frame(pos=1:1000, pvalinc=pval_inc, pvaldec=pval_dec)
  dfforvis_oddsratio <- data.frame(pos=1:1000, oddsratinc=oddsratio_inc, oddsratdec=oddsratio_dec)

  dfforvis$IncreasedEvents <- dfforvis$IncreasedEvents/rnamapfile_rbp_increased_incomp
  dfforvis$DecreasedEvents <- dfforvis$DecreasedEvents/rnamapfile_rbp_decreased_incomp
  dfforvis$MaintainedEvents <- dfforvis$MaintainedEvents/rnamapfile_rbp_maintained_incomp

  if (plot==TRUE){

    mapplot <- ggplot(dfforvis, aes(x = Name)) +
      geom_line(aes(y = IncreasedEvents, color = "IncreasedEvents"), size = 1) +
      geom_line(aes(y = DecreasedEvents, color = "DecreasedEvents"), size = 1) +
      geom_line(aes(y = MaintainedEvents, color = "MaintainedEvents"), size = 1) + xlim(0, 1000) +
      labs(title = paste0(rnaBP, " RNA Binding Map - ", title),
           x = "",
           y = " Normalised\nDensity\n") +
      scale_color_manual(values = c(IncreasedEvents = "#de425b", DecreasedEvents = "#769fca", MaintainedEvents = "black")) +
      theme_bw() +
      geom_vline(xintercept = c(0,50, 250, 450,500,550,750,950,1000), linetype = "dashed") +
      theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + theme(text=element_text(size=15))

    labels_plot <- ggplot() +
      annotate("text", x = 1, y = 0, label = paste0("Increased: ",length(unique(rnamapfile_rbp_increased_rbp$EVENT)), " (", nrow(ASfile_filt_increased), ")\n"),
               color = "#de425b", size = 4.5, fontface = "bold", hjust= -0.75) +
      annotate("text", x = 1, y = 0, label = paste0("Maintained: ",length(unique(rnamapfile_rbp_maintained_rbp$EVENT)), " (", nrow(ASfile_filt_maintained), ")\n"),
               color = "black", size = 4.5, fontface = "bold",hjust=0.5) +
      annotate("text", x = 1, y = 0, label = paste0("Decreased: ",length(unique(rnamapfile_rbp_decreased_rbp$EVENT)), " (", nrow(ASfile_filt_decreased), ")\n"),
               color = "#769fca", size = 4.5, fontface = "bold", hjust=1.75) +
      theme_void() +
      theme(plot.margin = margin(t = 0, r = 0, b = 0, l = 0))
  }

  # Shared y-axis limit helper. If ylim_metric is provided it must be a
  # length-2 numeric vector, e.g. c(-5, 5)
  metric_ylim_layer <- if (!is.null(ylim_metric)) {
    coord_cartesian(ylim = ylim_metric)   # coord_cartesian clips without dropping data
  } else {
    NULL
  }

  if (plot==TRUE){
    if (metric == "EffectSize"){
      metricplot <- ggplot(dfforvis_oddsratio, aes(x = pos)) +
        geom_line(aes(y = oddsratinc, color = "Increased"), size = 1) +
        geom_line(aes(y = oddsratdec, color = "Decreased"), size = 1) +
        xlim(0, 1000) +
        labs(x = "Position",
             y = "Chi-Squared") +
        theme_bw() +
        geom_vline(xintercept = c(0,50, 250, 450,500,550,750,950,1000), linetype = "dashed") +
        scale_color_manual(values = c("Increased" = "#de425b", "Decreased" = "#769fca")) +
        theme(axis.line = element_line(colour = "black"),
              panel.grid.minor = element_blank(),
              panel.border = element_blank(),
              panel.background = element_blank(),
              axis.text.x=element_blank(),
              axis.ticks.x=element_blank(),
              legend.position = "None") + theme(text=element_text(size=15)) +
        scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) +
        metric_ylim_layer
    }
    if (metric == "FDR"){
      metricplot <- ggplot(dfforvis_pval, aes(x = pos)) +
        geom_line(aes(y = pvalinc, color = "Increased"), size = 1) +
        geom_line(aes(y = pvaldec, color = "Decreased"), size = 1) +
        xlim(0, 1000) +
        labs(x = "Position",
             y = "-log10 (FDR)\n") +
        theme_bw() +
        geom_vline(xintercept = c(0,50, 250, 450,500,550,750,950,1000), linetype = "dashed") +
        scale_color_manual(values = c("Increased" = "#de425b", "Decreased" = "#769fca")) +
        theme(axis.line = element_line(colour = "black"),
              panel.grid.minor = element_blank(),
              panel.border = element_blank(),
              panel.background = element_blank(),
              axis.text.x=element_blank(),
              axis.ticks.x=element_blank(),
              legend.position = "None") + theme(text=element_text(size=15)) +
        scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) +
        metric_ylim_layer
    }

    completeplot <- ggarrange(mapplot, labels_plot, metricplot, ncol=1, heights = c(0.4,0.1,0.4), align="v")
  }

  if (plot){
    return(completeplot)
  } else {
    return(list(pval_data = dfforvis_pval, oddsratio_data = dfforvis_oddsratio))
  }
}
```

# `eCLIPSE_full_IR()`

Same shape as `eCLIPSE_full()` above, with the differences that follow directly from working on the intron-retention layout (see Section 7 of the tutorial): the position columns dropped before the per-position loop are `501`/`502` rather than `1001`/`1002` (`EVENTS`/`RBP` sit right after 500 position columns instead of 1000), the per-position loop runs `1:500` rather than `1:1000`, and the vertical guide lines drawn on the plots mark the intron-retention layout's 4 region boundaries (`0, 50, 250, 450, 500`) rather than the 8-region layout's (`0, 50, 250, 450, 500, 550, 750, 950, 1000`).

```{r}
eCLIPSE_full_IR <- function(rnamapfile, pvalthreshold=0.05, PSIthreshold=0.05, ASfile, rnaBP, metric = "FDR", plot = T, title = NULL){

  # --- FIX: was ASfile <- ASfile[grepl("HsaINT", Event.ID)] - see the note ---
  # --- at the top of this file. VastDB/VAST-TOOLS intron-retention IDs    ---
  # --- already look like "HsaINT...", so this change is a no-op for the   ---
  # --- existing VastDB IR path; it is here purely for consistency with    ---
  # --- eCLIPSE_full() above, in case this function is ever fed a          ---
  # --- differently-formatted IR table in the future.                     ---
  ASfile <- ASfile[ASfile$Event.ID %in% rnamapfile$EVENTS, ]
  AS_events <- ASfile$Event.ID

  ASfile_filt_increased <- subset(ASfile, dPSI>PSIthreshold)
  ASfile_filt_decreased <- subset(ASfile, dPSI< -PSIthreshold)
  ASfile_filt_maintained <- subset(ASfile, !(dPSI > PSIthreshold) & !(dPSI < -PSIthreshold))

  #And now the eCLIPSE file
  rnamapfile_rbp_increased <- subset(rnamapfile, EVENTS %in% ASfile_filt_increased$Event.ID)
  rnamapfile_rbp_maintained <- subset(rnamapfile, EVENTS %in% ASfile_filt_maintained$Event.ID)
  rnamapfile_rbp_decreased <- subset(rnamapfile, EVENTS %in% ASfile_filt_decreased$Event.ID)

  rnamapfile_rbp_increased_rbp <- subset(rnamapfile_rbp_increased, RBP==rnaBP)
  rnamapfile_rbp_increased_nrbp <- subset(rnamapfile_rbp_increased, RBP!=rnaBP)
  rnamapfile_rbp_increased_nrbp <- rnamapfile_rbp_increased_nrbp[!duplicated(rnamapfile_rbp_increased_nrbp$EVENTS), ]
  rnamapfile_rbp_increased_nrbp[rnamapfile_rbp_increased_nrbp == 1] <- 0
  rnamapfile_rbp_maintained_rbp <- subset(rnamapfile_rbp_maintained, RBP==rnaBP)
  rnamapfile_rbp_maintained_nrbp <-subset(rnamapfile_rbp_maintained, RBP!=rnaBP)
  rnamapfile_rbp_maintained_nrbp <- rnamapfile_rbp_maintained_nrbp[!duplicated(rnamapfile_rbp_maintained_nrbp$EVENTS), ]
  rnamapfile_rbp_maintained_nrbp[rnamapfile_rbp_maintained_nrbp == 1] <- 0
  rnamapfile_rbp_decreased_rbp <- subset(rnamapfile_rbp_decreased, RBP==rnaBP)
  rnamapfile_rbp_decreased_nrbp <- subset(rnamapfile_rbp_decreased, RBP!=rnaBP)
  rnamapfile_rbp_decreased_nrbp <- rnamapfile_rbp_decreased_nrbp[!duplicated(rnamapfile_rbp_decreased_nrbp$EVENTS), ]
  rnamapfile_rbp_decreased_nrbp[rnamapfile_rbp_decreased_nrbp == 1] <- 0

  rnamapfile_rbp_increased <- rbind(rnamapfile_rbp_increased_rbp, rnamapfile_rbp_increased_nrbp)
  rnamapfile_rbp_decreased <- rbind(rnamapfile_rbp_decreased_rbp, rnamapfile_rbp_decreased_nrbp)
  rnamapfile_rbp_maintained <- rbind(rnamapfile_rbp_maintained_rbp, rnamapfile_rbp_maintained_nrbp)

  if (nrow(rnamapfile_rbp_increased) < 10 ||
      nrow(rnamapfile_rbp_maintained) < 10 ||
      nrow(rnamapfile_rbp_decreased) < 10) {
    print(paste0(rnaBP," does not have enough events."))
    return()  # Skip to the next iteration
  }

  rnamapfile_rbp_increased <- rnamapfile_rbp_increased[,-c(501,502)]
  rnamapfile_rbp_increased <- apply(rnamapfile_rbp_increased, c(1, 2), as.numeric)
  rnamapfile_rbp_increased_comp <- colSums(rnamapfile_rbp_increased, na.rm = TRUE)
  rnamapfile_rbp_increased_incomp <- colSums(!is.na(rnamapfile_rbp_increased))

  rnamapfile_rbp_decreased <- rnamapfile_rbp_decreased[,-c(501,502)]
  rnamapfile_rbp_decreased <- apply(rnamapfile_rbp_decreased, c(1, 2), as.numeric)
  rnamapfile_rbp_decreased_comp <- colSums(rnamapfile_rbp_decreased, na.rm = TRUE)
  rnamapfile_rbp_decreased_incomp <- colSums(!is.na(rnamapfile_rbp_decreased))

  rnamapfile_rbp_maintained <- rnamapfile_rbp_maintained[,-c(501,502)]
  rnamapfile_rbp_maintained <- apply(rnamapfile_rbp_maintained, c(1, 2), as.numeric)
  rnamapfile_rbp_maintained_comp <- colSums(rnamapfile_rbp_maintained, na.rm = TRUE)
  rnamapfile_rbp_maintained_incomp <- colSums(!is.na(rnamapfile_rbp_maintained))

  #PLOTTING
  dfforvis <- data.frame(
    Name = as.numeric(names(rnamapfile_rbp_maintained_comp)),  # the names are the metagenomic coordinates
    IncreasedEvents = as.numeric(rnamapfile_rbp_increased_comp+1),
    DecreasedEvents = as.numeric(rnamapfile_rbp_decreased_comp+1),
    MaintainedEvents = as.numeric(rnamapfile_rbp_maintained_comp+1)
  )

  pval_inc <- c()
  pval_dec <- c()
  oddsratio_inc <- c()
  oddsratio_dec <- c()
  for (i in 1:500){
    inc_fish <- dfforvis[i,2]
    inc_fish_rev <- (rnamapfile_rbp_increased_incomp[i] - dfforvis[i,2])
    inc_fish_vs <- (dfforvis[i,4])
    inc_fish_vs_rev <- (rnamapfile_rbp_maintained_incomp[i])-(dfforvis[i,4])

    dec_fish <- dfforvis[i,3]
    dec_fish_rev <- (rnamapfile_rbp_decreased_incomp[i] - dfforvis[i,3])
    dec_fish_vs <- (dfforvis[i,4])
    dec_fish_vs_rev <- (rnamapfile_rbp_maintained_incomp[i])-(dfforvis[i,4])

    inc_df <- matrix(c(inc_fish, inc_fish_rev, inc_fish_vs, inc_fish_vs_rev), nrow = 2)
    inc_pval <- chisq.test(inc_df)$p.value
    inc_odds <- chisq.test(inc_df)$statistic

    dec_df <- matrix(c(dec_fish, dec_fish_rev, dec_fish_vs, dec_fish_vs_rev), nrow = 2)
    dec_pval <- chisq.test(dec_df)$p.value
    dec_odds <- chisq.test(dec_df)$statistic

    inc_pval <- -log10(p.adjust(inc_pval, method = "BH", n = 500))
    dec_pval <- -log10(p.adjust(dec_pval, method = "BH", n = 500))

    if ((inc_fish/rnamapfile_rbp_increased_incomp[i]) <=(inc_fish_vs/(rnamapfile_rbp_decreased_incomp[i]+rnamapfile_rbp_maintained_incomp[i]))){
      inc_pval <- inc_pval*-1
      inc_odds <- inc_odds*-1
    }
    if ((dec_fish/rnamapfile_rbp_decreased_incomp[i]) <=(dec_fish_vs/(rnamapfile_rbp_increased_incomp[i]+rnamapfile_rbp_maintained_incomp[i]))){
      dec_pval <- dec_pval*-1
      dec_odds <- dec_odds*-1
    }
    pval_inc <- c(pval_inc,inc_pval)
    oddsratio_inc <- c(oddsratio_inc, inc_odds)
    pval_dec <- c(pval_dec,dec_pval)
    oddsratio_dec <- c(oddsratio_dec, dec_odds)

  }

  dfforvis_pval <- data.frame(pos=1:500, pvalinc=pval_inc, pvaldec=pval_dec)
  dfforvis_oddsratio <- data.frame(pos=1:500, oddsratinc=oddsratio_inc, oddsratdec=oddsratio_dec)

  #The dataframe must be modified beforehand
  dfforvis$IncreasedEvents <- dfforvis$IncreasedEvents/rnamapfile_rbp_increased_incomp
  dfforvis$DecreasedEvents <- dfforvis$DecreasedEvents/rnamapfile_rbp_decreased_incomp
  dfforvis$MaintainedEvents <- dfforvis$MaintainedEvents/rnamapfile_rbp_maintained_incomp

  if (plot==TRUE){

    mapplot <- ggplot(dfforvis, aes(x = Name)) +
      geom_line(aes(y = IncreasedEvents, color = "IncreasedEvents"), size = 1) +
      geom_line(aes(y = DecreasedEvents, color = "DecreasedEvents"), size = 1) +
      geom_line(aes(y = MaintainedEvents, color = "MaintainedEvents"), size = 1) + xlim(0, 500) +
      labs(title = paste0(rnaBP, " RNA Binding Map - ", title),
           x = "",
           y = "Normalized\nDensity\n") +
      scale_color_manual(values = c(IncreasedEvents = "#de425b", DecreasedEvents = "#769fca", MaintainedEvents = "black")) +
      theme_bw() +
      geom_vline(xintercept = c(0,50, 250, 450,500), linetype = "dashed") +
      theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None")+theme(text=element_text(size=15))

    labels_plot <- ggplot() +
      annotate("text", x = 1, y = 0, label = paste0("Increased: ",length(table(rnamapfile_rbp_increased_rbp$EVENT)), " (", nrow(ASfile_filt_increased), ")\n"),
               color = "#de425b", size = 4, fontface = "bold", hjust= -0.75) +
      annotate("text", x = 1, y = 0, label = paste0("Maintained: ",length(table(rnamapfile_rbp_maintained_rbp$EVENT)), " (", nrow(ASfile_filt_maintained), ")\n"),
               color = "black", size = 4, fontface = "bold",hjust=0.5) +
      annotate("text", x = 1, y = 0, label = paste0("Decreased: ",length(table(rnamapfile_rbp_decreased_rbp$EVENT)), " (", nrow(ASfile_filt_decreased), ")\n"),
               color = "#769fca", size = 4, fontface = "bold", hjust=1.75) +
      theme_void() + # Removes all axes and background
      theme(plot.margin = margin(t = 0, r = 0, b = 0, l = 0))

  }

if (plot==TRUE){
    if (metric=="EffectSize"){
      metricplot <- ggplot(dfforvis_oddsratio, aes(x = pos)) +
        geom_line(aes(y = oddsratinc, color = "Increased"), size = 1) +
        geom_line(aes(y = oddsratdec, color = "Decreased"), size = 1) +
        xlim(0, 500) +
        labs(x = "Position",
             y = "Chi-Squared\nStatistic\n") +
        theme_bw() +
        geom_vline(xintercept = c(0, 50, 250, 450, 500), linetype = "dashed")+
        scale_color_manual(values = c("Increased" = "#de425b", "Decreased" = "#769fca"))+ theme(axis.line = element_line(colour = "black"),
                                                                                                panel.grid.minor = element_blank(),
                                                                                                panel.border = element_blank(),
                                                                                                panel.background = element_blank(),
                                                                                                axis.text.x=element_blank(),
                                                                                                axis.ticks.x=element_blank(),
                                                                                                legend.position = "None")+theme(text=element_text(size=15))+
        scale_y_continuous(
          labels = scales::label_number(accuracy = 0.01)
        )}
    if (metric=="FDR"){
      metricplot <- ggplot(dfforvis_pval, aes(x = pos)) +
        geom_line(aes(y = pvalinc, color = "Increased"), size = 1) +
        geom_line(aes(y = pvaldec, color = "Decreased"), size = 1) +
        xlim(0, 500) +
        labs(x = "Position",
             y = "-log10 (FDR)\n") +
        theme_bw() +
        geom_vline(xintercept = c(0, 50, 250, 450, 500), linetype = "dashed")+
        scale_color_manual(values = c("Increased" = "#de425b", "Decreased" = "#769fca"))+ theme(axis.line = element_line(colour = "black"),
                                                                                                panel.grid.minor = element_blank(),
                                                                                                panel.border = element_blank(),
                                                                                                panel.background = element_blank(),
                                                                                                axis.text.x=element_blank(),
                                                                                                axis.ticks.x=element_blank(),
                                                                                                legend.position = "None")+theme(text=element_text(size=15))+
        scale_y_continuous(
          labels = scales::label_number(accuracy = 0.01)
        )}

    completeplot <- ggarrange(mapplot, labels_plot, metricplot, ncol=1, heights = c(0.4,0.1,0.4), align="v")
  }

  if (plot){
    return(completeplot)
  }
  else{
    return(list(pval_data = dfforvis_pval, oddsratio_data = dfforvis_oddsratio))
  }
}
```

# Running it

```{r}
rnamapfile <- readRDS("~/path/to/your/project/eCLIPSE_position_matrix.rds")  # from Eclip_position_matrix.Rmd
ASfile     <- fread("~/path/to/your/project/my_betAS_table.txt")             # from "Dif Alternative Splicing Analysis.Rmd"

rnaBP <- "RBM39"
eCLIPSE_full(rnamapfile = rnamapfile, ASfile = ASfile, rnaBP = rnaBP,
             metric = "FDR", plot = TRUE, title = "control vs shRNA")
```

## Intron-retention path (VastDB only)

```{r}
rnamapfile_ir <- readRDS("~/path/to/your/project/eCLIPSE_position_matrix_IR.rds")
eCLIPSE_full_IR(rnamapfile = rnamapfile_ir, ASfile = ASfile, rnaBP = rnaBP,
                 metric = "FDR", plot = TRUE, title = "control vs shRNA, intron retention")
```

# Per-RBP binding heatmaps


```{r}
eCLIPSE_heatmap <- function(rnamapfile, ASfile, rbps, metric = "EffectSize",
                             pvalthreshold = 0.05, PSIthreshold = 0.05){

  # tryCatch, not just the NULL-filter below: an RBP name with real rows in
  # rnamapfile but too few events trips eCLIPSE_full()'s own "not enough
  # events" check and returns NULL cleanly, but an RBP name absent from
  # rnamapfile$RBP entirely skips that check (every row counts as "not this
  # RBP") and reaches chisq.test() with a malformed table instead, which
  # errors rather than returning NULL.
  results <- lapply(rbps, function(rbp) {
    tryCatch(
      eCLIPSE_full(rnamapfile = rnamapfile, pvalthreshold = pvalthreshold,
                   PSIthreshold = PSIthreshold, ASfile = ASfile, rnaBP = rbp,
                   metric = metric, plot = FALSE),
      error = function(e) NULL
    )
  })
  names(results) <- rbps
  results <- results[!vapply(results, is.null, logical(1))]
  if (length(results) == 0) stop("None of the requested RBPs had enough events.")

  data_slot <- if (metric == "EffectSize") "oddsratio_data" else "pval_data"
  inc_col   <- if (metric == "EffectSize") "oddsratinc" else "pvalinc"
  dec_col   <- if (metric == "EffectSize") "oddsratdec" else "pvaldec"

  inc_mat <- do.call(rbind, lapply(results, function(x) x[[data_slot]][[inc_col]]))
  dec_mat <- do.call(rbind, lapply(results, function(x) x[[data_slot]][[dec_col]]))
  rownames(inc_mat) <- rownames(dec_mat) <- names(results)

  # --- FIX: both metrics are signed (positive = enriched in the direction ---
  # --- being tested, negative = depleted), and the two panels' own value  ---
  # --- ranges can differ a lot. A colour scale fit independently per panel ---
  # --- (ggplot's default) makes the same colour mean different things in  ---
  # --- the decreased vs. increased panel. Use one symmetric scale (+/- the ---
  # --- largest absolute value across BOTH panels) for both instead, same  ---
  # --- as the CHARM app's own build_binding_heatmap_user() in             ---
  # --- helper_functions.R - which also applies to neither metric no log   ---
  # --- transform, so this replaces the signed-log1p transform used here   ---
  # --- previously.                                                        ---
  all_vals <- c(inc_mat, dec_mat)
  all_vals <- all_vals[is.finite(all_vals)]
  lim <- if (length(all_vals) == 0) c(-1, 1) else {
    mx <- max(abs(all_vals)); if (mx == 0) c(-1, 1) else c(-mx, mx)
  }

  fill_label <- if (metric == "EffectSize") "Chi-Squared\nStatistic\n" else "-log10(FDR)\n"
  boundaries <- c(0, 50, 250, 450, 500, 550, 750, 950, 1000)
  rbp_order  <- sort(rownames(inc_mat), decreasing = TRUE)

  melt_heat <- function(mat) {
    m <- reshape2::melt(mat)
    names(m) <- c("RBP", "Position", "value")
    m$RBP <- factor(m$RBP, levels = rbp_order)
    m
  }

  tile_plot <- function(mat, title){
    ggplot(melt_heat(mat), aes(x = Position, y = RBP, fill = value)) +
      geom_tile(color = "white") +
      scale_fill_gradient2(low = "#53A2BE", mid = "#EEEEEE", high = "#BA3B46",
                            name = fill_label, limits = lim, oob = scales::squish) +
      geom_vline(xintercept = boundaries, linetype = "dashed") +
      theme_minimal() +
      labs(x = "Position", y = NULL, title = title)
  }

  dec_plot <- tile_plot(dec_mat, "Decreased splicing") +
    theme(axis.text.y = element_blank(), legend.position = "none")
  inc_plot <- tile_plot(inc_mat, "Increased splicing")

  ggarrange(dec_plot, inc_plot, ncol = 2, widths = c(0.44, 0.56))
}
```

Same shape for the intron-retention layout: 4 region boundaries instead of 8, and the per-position vectors are 500 long instead of 1000.

```{r}
eCLIPSE_heatmap_IR <- function(rnamapfile, ASfile, rbps, metric = "EffectSize",
                                pvalthreshold = 0.05, PSIthreshold = 0.05){

  results <- lapply(rbps, function(rbp) {
    tryCatch(
      eCLIPSE_full_IR(rnamapfile = rnamapfile, pvalthreshold = pvalthreshold,
                       PSIthreshold = PSIthreshold, ASfile = ASfile, rnaBP = rbp,
                       metric = metric, plot = FALSE),
      error = function(e) NULL
    )
  })
  names(results) <- rbps
  results <- results[!vapply(results, is.null, logical(1))]
  if (length(results) == 0) stop("None of the requested RBPs had enough events.")

  data_slot <- if (metric == "EffectSize") "oddsratio_data" else "pval_data"
  inc_col   <- if (metric == "EffectSize") "oddsratinc" else "pvalinc"
  dec_col   <- if (metric == "EffectSize") "oddsratdec" else "pvaldec"

  inc_mat <- do.call(rbind, lapply(results, function(x) x[[data_slot]][[inc_col]]))
  dec_mat <- do.call(rbind, lapply(results, function(x) x[[data_slot]][[dec_col]]))
  rownames(inc_mat) <- rownames(dec_mat) <- names(results)

  all_vals <- c(inc_mat, dec_mat)
  all_vals <- all_vals[is.finite(all_vals)]
  lim <- if (length(all_vals) == 0) c(-1, 1) else {
    mx <- max(abs(all_vals)); if (mx == 0) c(-1, 1) else c(-mx, mx)
  }

  fill_label <- if (metric == "EffectSize") "Chi-Squared\nStatistic\n" else "-log10(FDR)\n"
  boundaries <- c(0, 50, 250, 450, 500)
  rbp_order  <- sort(rownames(inc_mat), decreasing = TRUE)

  melt_heat <- function(mat) {
    m <- reshape2::melt(mat)
    names(m) <- c("RBP", "Position", "value")
    m$RBP <- factor(m$RBP, levels = rbp_order)
    m
  }

  tile_plot <- function(mat, title){
    ggplot(melt_heat(mat), aes(x = Position, y = RBP, fill = value)) +
      geom_tile(color = "white") +
      scale_fill_gradient2(low = "#53A2BE", mid = "#EEEEEE", high = "#BA3B46",
                            name = fill_label, limits = lim, oob = scales::squish) +
      geom_vline(xintercept = boundaries, linetype = "dashed") +
      theme_minimal() +
      labs(x = "Position", y = NULL, title = title)
  }

  dec_plot <- tile_plot(dec_mat, "Decreased splicing") +
    theme(axis.text.y = element_blank(), legend.position = "none")
  inc_plot <- tile_plot(inc_mat, "Increased splicing")

  ggarrange(dec_plot, inc_plot, ncol = 2, widths = c(0.44, 0.56))
}
```

RBPs without enough events are dropped by `eCLIPSE_full()`/`eCLIPSE_full_IR()` themselves (with a printed message naming which ones), same as everywhere else these two functions are used.

## Running it

```{r}
rbps_of_interest <- c("SRSF1", "U2AF2")
eCLIPSE_heatmap(rnamapfile = rnamapfile, ASfile = ASfile,
                rbps = rbps_of_interest, metric = "EffectSize")

# Intron-retention path (VastDB only):
eCLIPSE_heatmap_IR(rnamapfile = rnamapfile_ir, ASfile = ASfile,
                    rbps = rbps_of_interest, metric = "EffectSize")
```

