---
title: "CHARM: Data Processing"
subtitle: "Comprehensive Hub for Alternative Regulatory Mapping"
author: 
  - name: "Alexandre Kaizeler^[Disease Transcriptomics Group, Gulbenkian Institute of Molecular Medicine]"
    email: alexandreafonso@gimm.pt
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  rmdformats::downcute:
    downcute_theme: "chaos"
---

# General

## Import libraries

```{r libraries, echo=FALSE}
# These are all the libraries I used throughout the analysis. Not all of them will be required on this guidebook
library(Rsubread)
library(limma)
library(edgeR)
library(DESeq2)
library(magrittr)
library(ggrepel)
library(tidyverse)
library(vsn)
library(ggplot2)
library(NMF)
library(biomaRt)
library(ggpubr)
library(factoextra)
library(ggfortify)
library(plotly)
library(reshape2)
library(fgsea)
library(msigdbr)
library(data.table)
library(colorspace)
library(readxl)
library(fs)
library(betAS)
library(Matrix)
library(pheatmap)
library(Rtsne)
set.seed(1906)
```

# A. Data Processing 

*Date: August 22th 2025*

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

### Import data

To create the R object that we will use for CHARM, we need three data types:

Expression: Differential expression upon RBP knockdown/knockout, with pathway-level insights from Gene Set Enrichment Analysis.

Splicing: Alternative splicing changes quantified with betAS (Ferreira et al., 2024), using VastDB nomenclature (Tapial et al., 2017).

Binding: Altered RNA binding patterns of the silenced RBP and other RBPs, characterized with the eCLIPSE tool.

### Binding

This is a complex process, so we will start by simplifying the data we already have. First, we sabe all the tables from betAS has .txt files. This is because since this process is very heavy and slow, we won't use R for this, but instead python. To facilitate file usage between the two programming languages, these files will always be saved as .txt or other similar file types.

```{r}

CharmObj_rbp_All <- readRDS("~/Projects/CHARM/data/Charm.object.RDS")
CharmObj_rbp_HEPG2 <- readRDS("~/Projects/CHARM/data/Charm.object_HEPG2.RDS")
CharmObj_rbp_K562 <- readRDS("~/Projects/CHARM/data/Charm.object_K562.RDS")
eCLIPSE_BigExon_full <- fread("~/Projects/StressGranules/AS.WC_Transcriptome/eCLIPSE/eCLIPSE_BigExon.csv", header = T)[,-1]

for (rbp in unique(eCLIPSE_BigExon_full$RBP)) {
  if (!is.null(CharmObj_rbp_All[[rbp]])) {
    betAsframe <- CharmObj_rbp_All[[rbp]]$VulcanTable
    
    outfile <- file.path(
      path.expand("~/Projects/StressGranules/AS.WC_Transcriptome"),
      paste0("shRNAExp/", rbp,"/", rbp,  "_VulcanTable_both.txt")
    )
    
    write.table(betAsframe, outfile, row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)
    message("Saved: ", outfile)
  } else {
    message("Skipping missing RBP: ", rbp)
  }
}

for (rbp in unique(eCLIPSE_BigExon_full$RBP)) {
  if (!is.null(CharmObj_rbp_HEPG2[[rbp]])) {
    betAsframe <- CharmObj_rbp_HEPG2[[rbp]]$VulcanTable
    
    outfile <- file.path(
      path.expand("~/Projects/StressGranules/AS.WC_Transcriptome"),
      paste0("shRNAExp/", rbp,"/", rbp,  "_VulcanTable_HEPG2.txt")
    )
    
    write.table(betAsframe, outfile, row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)
    message("Saved: ", outfile)
  } else {
    message("Skipping missing RBP: ", rbp)
  }
}

for (rbp in unique(eCLIPSE_BigExon_full$RBP)) {
  if (!is.null(CharmObj_rbp_K562[[rbp]])) {
    betAsframe <- CharmObj_rbp_K562[[rbp]]$VulcanTable
    
    outfile <- file.path(
      path.expand("~/Projects/StressGranules/AS.WC_Transcriptome"),
      paste0("shRNAExp/", rbp,"/", rbp,  "_VulcanTable_K562.txt")
    )
    
    write.table(betAsframe, outfile, row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)
    message("Saved: ", outfile)
  } else {
    message("Skipping missing RBP: ", rbp)
  }
}


```

### File Filtration

```{r}
EventsBoth <- readRDS("~/Projects/CHARM/data/AvailableEvents_Both.RDS")
EventsK562 <- readRDS("~/Projects/CHARM/data/AvailableEvents_K562.RDS")
EventsHEPG2 <- readRDS("~/Projects/CHARM/data/AvailableEvents_HEPG2.RDS")

eCLIPSE_BigExon_full_both <- eCLIPSE_BigExon_full[eCLIPSE_BigExon_full$EVENTS %in% EventsBoth,]
eCLIPSE_BigExon_full_K562 <- eCLIPSE_BigExon_full[eCLIPSE_BigExon_full$EVENTS %in% EventsK562,]
eCLIPSE_BigExon_full_HEPG2 <- eCLIPSE_BigExon_full[eCLIPSE_BigExon_full$EVENTS %in% EventsHEPG2,]

write.table(eCLIPSE_BigExon_full_both, "~/Projects/CHARM/data/eCLIPSE_BigExon_both.txt", row.names = F, col.names = T)
write.table(eCLIPSE_BigExon_full_K562, "~/Projects/CHARM/data/eCLIPSE_BigExon_K562.txt", row.names = F, col.names = T)
write.table(eCLIPSE_BigExon_full_HEPG2, "~/Projects/CHARM/data/eCLIPSE_BigExon_HEPG2.txt", row.names = F, col.names = T)

eCLIPSE_Intron_full_both <- eCLIPSE_Intron_full[eCLIPSE_Intron_full$EVENTS %in% EventsBoth,]
eCLIPSE_Intron_full_K562 <- eCLIPSE_Intron_full[eCLIPSE_Intron_full$EVENTS %in% EventsK562,]
eCLIPSE_Intron_full_HEPG2 <- eCLIPSE_Intron_full[eCLIPSE_Intron_full$EVENTS %in% EventsHEPG2,]

write.table(eCLIPSE_Intron_full_both, "~/Projects/CHARM/data/eCLIPSE_Intron_both.txt", row.names = F, col.names = T)
write.table(eCLIPSE_Intron_full_K562, "~/Projects/CHARM/data/eCLIPSE_Intron_K562.txt", row.names = F, col.names = T)
write.table(eCLIPSE_Intron_full_HEPG2, "~/Projects/CHARM/data/eCLIPSE_Intron_HEPG2.txt", row.names = F, col.names = T)
```


# B. Function preparation (R Based)

These are the relevant functions I have used as a standalone eCLIPSE tool. Here, I will make some changes to accomodate and simplyfy the process.

## Exon Skipping Analysis

```{r}
eCLIPSE_BigExon_full <- fread("~/Projects/StressGranules/AS.WC_Transcriptome/eCLIPSE/eCLIPSE_BigExon.csv", header = T)[,-1]


eCLIPSE_heatmap <- function(metric, rnamapfile, pvalthreshold=0.05, PSIthreshold=0.05, ASfile, rnaBP, plot=FALSE){
  
  ASfile <- ASfile[grepl("HsaEX", Event.ID)]
  
  AS_events <- ASfile$Event.ID
 
  ASfile_filt_increased <- subset(ASfile, dPSI>PSIthreshold)
  ASfile_filt_decreased <- subset(ASfile, dPSI< -PSIthreshold)
  ASfile_filt_maintained <- subset(ASfile, !(dPSI > PSIthreshold) & !(dPSI < -PSIthreshold))

  #And now the eCLIPSE file
  rnamapfile_rbp_increased <- subset(rnamapfile, EVENTS %in% ASfile_filt_increased$Event.ID)
  rnamapfile_rbp_maintained <- subset(rnamapfile, EVENTS %in% ASfile_filt_maintained$Event.ID)
  rnamapfile_rbp_decreased <- subset(rnamapfile, EVENTS %in% ASfile_filt_decreased$Event.ID)
  
  rnamapfile_rbp_increased_rbp <- subset(rnamapfile_rbp_increased, RBP==rnaBP)
  rnamapfile_rbp_increased_nrbp <- subset(rnamapfile_rbp_increased, RBP!=rnaBP)
  rnamapfile_rbp_increased_nrbp <- rnamapfile_rbp_increased_nrbp[!duplicated(rnamapfile_rbp_increased_nrbp$EVENTS), ]
  rnamapfile_rbp_increased_nrbp[rnamapfile_rbp_increased_nrbp == 1] <- 0
  rnamapfile_rbp_maintained_rbp <- subset(rnamapfile_rbp_maintained, RBP==rnaBP)
  rnamapfile_rbp_maintained_nrbp <-subset(rnamapfile_rbp_maintained, RBP!=rnaBP)
  rnamapfile_rbp_maintained_nrbp <- rnamapfile_rbp_maintained_nrbp[!duplicated(rnamapfile_rbp_maintained_nrbp$EVENTS), ]
  rnamapfile_rbp_maintained_nrbp[rnamapfile_rbp_maintained_nrbp == 1] <- 0
  rnamapfile_rbp_decreased_rbp <- subset(rnamapfile_rbp_decreased, RBP==rnaBP)
  rnamapfile_rbp_decreased_nrbp <- subset(rnamapfile_rbp_decreased, RBP!=rnaBP)
  rnamapfile_rbp_decreased_nrbp <- rnamapfile_rbp_decreased_nrbp[!duplicated(rnamapfile_rbp_decreased_nrbp$EVENTS), ]
  rnamapfile_rbp_decreased_nrbp[rnamapfile_rbp_decreased_nrbp == 1] <- 0
  
  rnamapfile_rbp_increased <- rbind(rnamapfile_rbp_increased_rbp, rnamapfile_rbp_increased_nrbp)
  rnamapfile_rbp_decreased <- rbind(rnamapfile_rbp_decreased_rbp, rnamapfile_rbp_decreased_nrbp)
  rnamapfile_rbp_maintained <- rbind(rnamapfile_rbp_maintained_rbp, rnamapfile_rbp_maintained_nrbp)
  
  if (nrow(rnamapfile_rbp_increased) < 10 || 
      nrow(rnamapfile_rbp_maintained) < 10 || 
      nrow(rnamapfile_rbp_decreased) < 10) {
    print(paste0(rnaBP," does not have enough events."))
    return()  # Skip to the next iteration
  }
  
  rnamapfile_rbp_increased <- rnamapfile_rbp_increased[,-c(1001,1002)]
  rnamapfile_rbp_increased <- apply(rnamapfile_rbp_increased, c(1, 2), as.numeric)
  rnamapfile_rbp_increased_comp <- colSums(rnamapfile_rbp_increased, na.rm = TRUE)
  rnamapfile_rbp_increased_incomp <- colSums(!is.na(rnamapfile_rbp_increased))
  
  rnamapfile_rbp_decreased <- rnamapfile_rbp_decreased[,-c(1001,1002)]
  rnamapfile_rbp_decreased <- apply(rnamapfile_rbp_decreased, c(1, 2), as.numeric)
  rnamapfile_rbp_decreased_comp <- colSums(rnamapfile_rbp_decreased, na.rm = TRUE)
  rnamapfile_rbp_decreased_incomp <- colSums(!is.na(rnamapfile_rbp_decreased))
  
  
  rnamapfile_rbp_maintained <- rnamapfile_rbp_maintained[,-c(1001,1002)]
  rnamapfile_rbp_maintained <- apply(rnamapfile_rbp_maintained, c(1, 2), as.numeric)
  rnamapfile_rbp_maintained_comp <- colSums(rnamapfile_rbp_maintained, na.rm = TRUE)
  rnamapfile_rbp_maintained_incomp <- colSums(!is.na(rnamapfile_rbp_maintained))

  #PLOTTING
  dfforvis <- data.frame(
    Name = as.numeric(names(rnamapfile_rbp_maintained_comp)),  # the names are the metagenomic coordinates
    IncreasedEvents = as.numeric(rnamapfile_rbp_increased_comp+1),
    DecreasedEvents = as.numeric(rnamapfile_rbp_decreased_comp+1),
    MaintainedEvents = as.numeric(rnamapfile_rbp_maintained_comp+1)
  )
  
  
  pval_inc <- c()
  pval_dec <- c()
  oddsratio_inc <- c()
  oddsratio_dec <- c()
  for (i in 1:1000){
    inc_fish <- dfforvis[i,2]
    inc_fish_rev <- (rnamapfile_rbp_increased_incomp[i] - dfforvis[i,2])
    inc_fish_vs <- (dfforvis[i,4])
    inc_fish_vs_rev <- (rnamapfile_rbp_maintained_incomp[i])-(dfforvis[i,4])
    
    dec_fish <- dfforvis[i,3]
    dec_fish_rev <- (rnamapfile_rbp_decreased_incomp[i] - dfforvis[i,3])
    dec_fish_vs <- (dfforvis[i,4])
    dec_fish_vs_rev <- (rnamapfile_rbp_maintained_incomp[i])-(dfforvis[i,4])
    
    inc_df <- matrix(c(inc_fish, inc_fish_rev, inc_fish_vs, inc_fish_vs_rev), nrow = 2)
    inc_pval <- chisq.test(inc_df)$p.value
    inc_odds <- chisq.test(inc_df)$statistic
    
    dec_df <- matrix(c(dec_fish, dec_fish_rev, dec_fish_vs, dec_fish_vs_rev), nrow = 2)
    dec_pval <- chisq.test(dec_df)$p.value
    dec_odds <- chisq.test(dec_df)$statistic
    
    inc_pval <- -log10(p.adjust(inc_pval, method = "BH", n = 1000))
    dec_pval <- -log10(p.adjust(dec_pval, method = "BH", n = 1000))
    
    if ((inc_fish/rnamapfile_rbp_increased_incomp[i]) <=(inc_fish_vs/(rnamapfile_rbp_maintained_incomp[i]))){
      inc_pval <- inc_pval*-1
      inc_odds <- inc_odds*-1
    }
    if ((dec_fish/rnamapfile_rbp_decreased_incomp[i]) <=(dec_fish_vs/(rnamapfile_rbp_maintained_incomp[i]))){
      dec_pval <- dec_pval*-1
      dec_odds <- dec_odds*-1
    }
    pval_inc <- c(pval_inc,inc_pval)
    oddsratio_inc <- c(oddsratio_inc, inc_odds)
    pval_dec <- c(pval_dec,dec_pval)
    oddsratio_dec <- c(oddsratio_dec, dec_odds)
    
  }
  
dfforvis_pval <- data.frame(pos = 1:1000, pvalinc = pval_inc, pvaldec = pval_dec)
dfforvis_oddsratio <- data.frame(pos = 1:1000, oddsratinc = oddsratio_inc, oddsratdec = oddsratio_dec)

# Return both data frames as a list, regardless of the 'metric'
return(list(pval_data = dfforvis_pval, oddsratio_data = dfforvis_oddsratio))
}






# Initialize results list
eCLIPSE_heatmap_results <- list()

rbps <- unique(eCLIPSE_BigExon_full$RBP)
targets <- rbps

for (rbp in rbps) {
  file_path <- file.path("~/Projects/StressGranules/AS.WC_Transcriptome/shRNAExp", rbp, paste0(rbp, "_VulcanTable_both.txt"))
  
  if (!file.exists(file_path)) {
    message("Skipping missing file for ", rbp)
    next
  }
  
  message("Processing ", rbp, " ...")
  betAS <- fread(file_path, header = TRUE)
  eCLIPSE_heatmap_results[[rbp]] <- list()
  
  # Initialize collector for saving later
  all_metrics <- list()
  
  for (target in targets) {
    message("  Target: ", target)
    
    dpsi_values <- seq(0.01, 0.1, by = 0.01)
    temp_results <- list()
    
    # Run eCLIPSE_heatmap for multiple dPSIs
    for (dpsi in dpsi_values) {
      res <- tryCatch({
        eCLIPSE_heatmap(
          metric = NULL,
          rnamapfile = eCLIPSE_BigExon_full,
          pvalthreshold = 0.05,
          PSIthreshold = dpsi,
          ASfile = betAS,
          rnaBP = target,
          plot = FALSE
        )
      }, error = function(e) {
        message("    Error for dPSI=", dpsi, ": ", e$message)
        NULL
      })
      
      if (!is.null(res)) {
        temp_results[[as.character(dpsi)]] <- res
      }
    }
    
    if (length(temp_results) == 0) next
    
    # Compute metrics summary
    stats <- data.frame(
      dpsi = as.numeric(names(temp_results)),
      sum_pvalinc = sapply(temp_results, function(x) sum(x$pval_data$pvalinc, na.rm = TRUE)),
      sum_pvaldec = sapply(temp_results, function(x) sum(x$pval_data$pvaldec, na.rm = TRUE)),
      sum_oddsinc = sapply(temp_results, function(x) sum(x$oddsratio_data$oddsratinc, na.rm = TRUE)),
      sum_oddsdec = sapply(temp_results, function(x) sum(x$oddsratio_data$oddsratdec, na.rm = TRUE))
    )
    
    # Pick relevant dPSIs (fixed + best)
    best_pvalinc <- stats$dpsi[which.max(stats$sum_pvalinc)]
    best_pvaldec <- stats$dpsi[which.max(stats$sum_pvaldec)]
    best_oddsinc <- stats$dpsi[which.max(stats$sum_oddsinc)]
    best_oddsdec <- stats$dpsi[which.max(stats$sum_oddsdec)]
    
    keep_dpsis <- unique(na.omit(c(0.05, 0.1, best_pvalinc, best_pvaldec, best_oddsinc, best_oddsdec)))
    message("    Keeping dPSIs: ", paste(keep_dpsis, collapse = ", "))
    
    # Save results to list
    eCLIPSE_heatmap_results[[rbp]][[target]] <- temp_results[as.character(keep_dpsis)]
    
    # Flatten into tidy table
    for (dpsi_sel in keep_dpsis) {
      res <- temp_results[[as.character(dpsi_sel)]]
      if (is.null(res)) next
      
      df_long <- rbind(
        data.frame(Pos = res$pval_data$pos, Metric = "pvalinc", Value = res$pval_data$pvalinc, dPSI = dpsi_sel, Target = target),
        data.frame(Pos = res$pval_data$pos, Metric = "pvaldec", Value = res$pval_data$pvaldec, dPSI = dpsi_sel, Target = target),
        data.frame(Pos = res$oddsratio_data$pos, Metric = "oddsratinc", Value = res$oddsratio_data$oddsratinc, dPSI = dpsi_sel, Target = target),
        data.frame(Pos = res$oddsratio_data$pos, Metric = "oddsratdec", Value = res$oddsratio_data$oddsratdec, dPSI = dpsi_sel, Target = target)
      )
      
      all_metrics[[length(all_metrics) + 1]] <- df_long
    }
  }
  
  # Save all results for this RBP as a .txt file
  if (length(all_metrics) > 0) {
    combined_df <- rbindlist(all_metrics)
    
    outfile <- file.path(
      path.expand("~/Projects/StressGranules/AS.WC_Transcriptome"),
      paste0("shRNAExp/", rbp, "/", rbp, "_BindingValues_both.txt")
    )
    
    fwrite(combined_df, outfile, sep = "\t")
    message("✅ Saved ", nrow(combined_df), " rows for ", rbp, " → ", outfile)
  } else {
    message("⚠️ No results to save for ", rbp)
  }
}


```

## Intron Retention Analysis

```{r}
eCLIPSE_Intron_full <- fread("~/Projects/StressGranules/AS.WC_Transcriptome/eCLIPSE/eCLIPSE_INTRONRET.txt", header = T)


eCLIPSE_heatmap_IR <- function(metric, rnamapfile, pvalthreshold=0.05, PSIthreshold=0.05, ASfile, rnaBP, plot=FALSE){
  
  ASfile <- ASfile[grepl("HsaINT", Event.ID)]


  AS_events <- ASfile$Event.ID
 
  ASfile_filt_increased <- subset(ASfile, dPSI>PSIthreshold)
  ASfile_filt_decreased <- subset(ASfile, dPSI< -PSIthreshold)
  ASfile_filt_maintained <- subset(ASfile, !(dPSI > PSIthreshold) & !(dPSI < -PSIthreshold))

  #And now the eCLIPSE file
  rnamapfile_rbp_increased <- subset(rnamapfile, EVENTS %in% ASfile_filt_increased$Event.ID)
  rnamapfile_rbp_maintained <- subset(rnamapfile, EVENTS %in% ASfile_filt_maintained$Event.ID)
  rnamapfile_rbp_decreased <- subset(rnamapfile, EVENTS %in% ASfile_filt_decreased$Event.ID)
  
  rnamapfile_rbp_increased_rbp <- subset(rnamapfile_rbp_increased, RBP==rnaBP)
  rnamapfile_rbp_increased_nrbp <- subset(rnamapfile_rbp_increased, RBP!=rnaBP)
  rnamapfile_rbp_increased_nrbp <- rnamapfile_rbp_increased_nrbp[!duplicated(rnamapfile_rbp_increased_nrbp$EVENTS), ]
  rnamapfile_rbp_increased_nrbp[rnamapfile_rbp_increased_nrbp == 1] <- 0
  rnamapfile_rbp_maintained_rbp <- subset(rnamapfile_rbp_maintained, RBP==rnaBP)
  rnamapfile_rbp_maintained_nrbp <-subset(rnamapfile_rbp_maintained, RBP!=rnaBP)
  rnamapfile_rbp_maintained_nrbp <- rnamapfile_rbp_maintained_nrbp[!duplicated(rnamapfile_rbp_maintained_nrbp$EVENTS), ]
  rnamapfile_rbp_maintained_nrbp[rnamapfile_rbp_maintained_nrbp == 1] <- 0
  rnamapfile_rbp_decreased_rbp <- subset(rnamapfile_rbp_decreased, RBP==rnaBP)
  rnamapfile_rbp_decreased_nrbp <- subset(rnamapfile_rbp_decreased, RBP!=rnaBP)
  rnamapfile_rbp_decreased_nrbp <- rnamapfile_rbp_decreased_nrbp[!duplicated(rnamapfile_rbp_decreased_nrbp$EVENTS), ]
  rnamapfile_rbp_decreased_nrbp[rnamapfile_rbp_decreased_nrbp == 1] <- 0
  
  rnamapfile_rbp_increased <- rbind(rnamapfile_rbp_increased_rbp, rnamapfile_rbp_increased_nrbp)
  rnamapfile_rbp_decreased <- rbind(rnamapfile_rbp_decreased_rbp, rnamapfile_rbp_decreased_nrbp)
  rnamapfile_rbp_maintained <- rbind(rnamapfile_rbp_maintained_rbp, rnamapfile_rbp_maintained_nrbp)
  
  if (nrow(rnamapfile_rbp_increased) < 10 || 
      nrow(rnamapfile_rbp_maintained) < 10 || 
      nrow(rnamapfile_rbp_decreased) < 10) {
    print(paste0(rnaBP," does not have enough events."))
    return()  # Skip to the next iteration
  }
  
  rnamapfile_rbp_increased <- rnamapfile_rbp_increased[,-c(501,502)]
  rnamapfile_rbp_increased <- apply(rnamapfile_rbp_increased, c(1, 2), as.numeric)
  rnamapfile_rbp_increased_comp <- colSums(rnamapfile_rbp_increased, na.rm = TRUE)
  rnamapfile_rbp_increased_incomp <- colSums(!is.na(rnamapfile_rbp_increased))
  
  rnamapfile_rbp_decreased <- rnamapfile_rbp_decreased[,-c(501,502)]
  rnamapfile_rbp_decreased <- apply(rnamapfile_rbp_decreased, c(1, 2), as.numeric)
  rnamapfile_rbp_decreased_comp <- colSums(rnamapfile_rbp_decreased, na.rm = TRUE)
  rnamapfile_rbp_decreased_incomp <- colSums(!is.na(rnamapfile_rbp_decreased))
  
  
  rnamapfile_rbp_maintained <- rnamapfile_rbp_maintained[,-c(501,502)]
  rnamapfile_rbp_maintained <- apply(rnamapfile_rbp_maintained, c(1, 2), as.numeric)
  rnamapfile_rbp_maintained_comp <- colSums(rnamapfile_rbp_maintained, na.rm = TRUE)
  rnamapfile_rbp_maintained_incomp <- colSums(!is.na(rnamapfile_rbp_maintained))

  #PLOTTING
  dfforvis <- data.frame(
    Name = as.numeric(names(rnamapfile_rbp_maintained_comp)),  # the names are the metagenomic coordinates
    IncreasedEvents = as.numeric(rnamapfile_rbp_increased_comp+1),
    DecreasedEvents = as.numeric(rnamapfile_rbp_decreased_comp+1),
    MaintainedEvents = as.numeric(rnamapfile_rbp_maintained_comp+1)
  )
  
  
  pval_inc <- c()
  pval_dec <- c()
  oddsratio_inc <- c()
  oddsratio_dec <- c()
  for (i in 1:500){
    inc_fish <- dfforvis[i,2]
    inc_fish_rev <- (rnamapfile_rbp_increased_incomp[i] - dfforvis[i,2])
    inc_fish_vs <- (dfforvis[i,4])
    inc_fish_vs_rev <- (rnamapfile_rbp_maintained_incomp[i])-(dfforvis[i,4])
    
    dec_fish <- dfforvis[i,3]
    dec_fish_rev <- (rnamapfile_rbp_decreased_incomp[i] - dfforvis[i,3])
    dec_fish_vs <- (dfforvis[i,4])
    dec_fish_vs_rev <- (rnamapfile_rbp_maintained_incomp[i])-(dfforvis[i,4])
    
    inc_df <- matrix(c(inc_fish, inc_fish_rev, inc_fish_vs, inc_fish_vs_rev), nrow = 2)
    inc_pval <- chisq.test(inc_df)$p.value
    inc_odds <- chisq.test(inc_df)$statistic
    
    dec_df <- matrix(c(dec_fish, dec_fish_rev, dec_fish_vs, dec_fish_vs_rev), nrow = 2)
    dec_pval <- chisq.test(dec_df)$p.value
    dec_odds <- chisq.test(dec_df)$statistic
    
    inc_pval <- -log10(p.adjust(inc_pval, method = "BH", n = 500))
    dec_pval <- -log10(p.adjust(dec_pval, method = "BH", n = 500))
    
    if ((inc_fish/rnamapfile_rbp_increased_incomp[i]) <=(inc_fish_vs/(rnamapfile_rbp_decreased_incomp[i]+rnamapfile_rbp_maintained_incomp[i]))){
      inc_pval <- inc_pval*-1
      inc_odds <- inc_odds*-1
    }
    if ((dec_fish/rnamapfile_rbp_decreased_incomp[i]) <=(dec_fish_vs/(rnamapfile_rbp_increased_incomp[i]+rnamapfile_rbp_maintained_incomp[i]))){
      dec_pval <- dec_pval*-1
      dec_odds <- dec_odds*-1
      
    }
    pval_inc <- c(pval_inc,inc_pval)
    oddsratio_inc <- c(oddsratio_inc, inc_odds)
    pval_dec <- c(pval_dec,dec_pval)
    oddsratio_dec <- c(oddsratio_dec, dec_odds)
    
  }
  
dfforvis_pval <- data.frame(pos = 1:500, pvalinc = pval_inc, pvaldec = pval_dec)
dfforvis_oddsratio <- data.frame(pos = 1:500, oddsratinc = oddsratio_inc, oddsratdec = oddsratio_dec)

# Return both data frames as a list, regardless of the 'metric'
return(list(pval_data = dfforvis_pval, oddsratio_data = dfforvis_oddsratio))
}


# Initialize results list
eCLIPSE_heatmap_results <- list()

rbps <- unique(eCLIPSE_Intron_full$RBP)
targets <- rbps

for (rbp in rbps) {
  file_path <- file.path("~/Projects/StressGranules/AS.WC_Transcriptome/shRNAExp", rbp, paste0(rbp, "_VulcanTable_both.txt"))
  
  if (!file.exists(file_path)) {
    message("Skipping missing file for ", rbp)
    next
  }
  
  message("Processing ", rbp, " ...")
  betAS <- fread(file_path, header = TRUE)
  eCLIPSE_heatmap_results[[rbp]] <- list()
  
  # Initialize collector for saving later
  all_metrics <- list()
  
  for (target in targets) {
    message("  Target: ", target)
    
    dpsi_values <- seq(0.01, 0.1, by = 0.01)
    temp_results <- list()
    
    # Run eCLIPSE_heatmap for multiple dPSIs
    for (dpsi in dpsi_values) {
      res <- tryCatch({
        eCLIPSE_heatmap_IR(
          metric = NULL,
          rnamapfile = eCLIPSE_Intron_full,
          pvalthreshold = 0.05,
          PSIthreshold = dpsi,
          ASfile = betAS,
          rnaBP = target,
          plot = FALSE
        )
      }, error = function(e) {
        message("    Error for dPSI=", dpsi, ": ", e$message)
        NULL
      })
      
      if (!is.null(res)) {
        temp_results[[as.character(dpsi)]] <- res
      }
    }
    
    if (length(temp_results) == 0) next
    
    # Compute metrics summary
    stats <- data.frame(
      dpsi = as.numeric(names(temp_results)),
      sum_pvalinc = sapply(temp_results, function(x) sum(x$pval_data$pvalinc, na.rm = TRUE)),
      sum_pvaldec = sapply(temp_results, function(x) sum(x$pval_data$pvaldec, na.rm = TRUE)),
      sum_oddsinc = sapply(temp_results, function(x) sum(x$oddsratio_data$oddsratinc, na.rm = TRUE)),
      sum_oddsdec = sapply(temp_results, function(x) sum(x$oddsratio_data$oddsratdec, na.rm = TRUE))
    )
    
    # Pick relevant dPSIs (fixed + best)
    best_pvalinc <- stats$dpsi[which.max(stats$sum_pvalinc)]
    best_pvaldec <- stats$dpsi[which.max(stats$sum_pvaldec)]
    best_oddsinc <- stats$dpsi[which.max(stats$sum_oddsinc)]
    best_oddsdec <- stats$dpsi[which.max(stats$sum_oddsdec)]
    
    keep_dpsis <- unique(na.omit(c(0.05, 0.1, best_pvalinc, best_pvaldec, best_oddsinc, best_oddsdec)))
    message("    Keeping dPSIs: ", paste(keep_dpsis, collapse = ", "))
    
    # Save results to list
    eCLIPSE_heatmap_results[[rbp]][[target]] <- temp_results[as.character(keep_dpsis)]
    
    # Flatten into tidy table
    for (dpsi_sel in keep_dpsis) {
      res <- temp_results[[as.character(dpsi_sel)]]
      if (is.null(res)) next
      
      df_long <- rbind(
        data.frame(Pos = res$pval_data$pos, Metric = "pvalinc", Value = res$pval_data$pvalinc, dPSI = dpsi_sel, Target = target),
        data.frame(Pos = res$pval_data$pos, Metric = "pvaldec", Value = res$pval_data$pvaldec, dPSI = dpsi_sel, Target = target),
        data.frame(Pos = res$oddsratio_data$pos, Metric = "oddsratinc", Value = res$oddsratio_data$oddsratinc, dPSI = dpsi_sel, Target = target),
        data.frame(Pos = res$oddsratio_data$pos, Metric = "oddsratdec", Value = res$oddsratio_data$oddsratdec, dPSI = dpsi_sel, Target = target)
      )
      
      all_metrics[[length(all_metrics) + 1]] <- df_long
    }
  }
  
  # Save all results for this RBP as a .txt file
  if (length(all_metrics) > 0) {
    combined_df <- rbindlist(all_metrics)
    
    outfile <- file.path(
      path.expand("~/Projects/StressGranules/AS.WC_Transcriptome"),
      paste0("shRNAExp/", rbp, "/", rbp, "_BindingValues_IR_both.txt")
    )
    
    fwrite(combined_df, outfile, sep = "\t")
    message("✅ Saved ", nrow(combined_df), " rows for ", rbp, " → ", outfile)
  } else {
    message("⚠️ No results to save for ", rbp)
  }
}

```

# C. Results Integration

## Exon Skipping

```{r}
build_bindingvalues_list <- function(
  base_path = "~/Projects/StressGranules/AS.WC_Transcriptome/shRNAExp"
) {
  # Expand base path
  base_path <- fs::path_expand(base_path)

  # List all immediate subdirectories (each RBP)
  subdirs <- fs::dir_ls(base_path, type = "directory", recurse = FALSE)
  folder_names <- fs::path_file(subdirs)

  # Initialize list for results
  out_list <- list()

  # Iterate through each folder and look for BindingValues_both.txt
  for (folder in folder_names) {
    folder_path <- fs::path(base_path, folder)
    
    # Find BindingValues_both.txt file
    matches <- fs::dir_ls(folder_path, regexp = "BindingValues_both\\.txt$", type = "file")
    
    if (length(matches) == 0) {
      message("No 'BindingValues_both.txt' file found in: ", folder_path)
      next  # skip this folder entirely
    }
    
    if (length(matches) > 1) {
      info <- fs::file_info(matches)
      matches <- matches[order(info$modification_time, decreasing = TRUE)]
      message("Multiple BindingValues_both.txt files in ", folder_path, 
              "; using latest: ", fs::path_file(matches[1]))
    }

    # Read the selected file
    dt <- data.table::fread(matches[1])

    # Add to output list, using folder name as key
    out_list[[folder]] <- list(bindingvalues = dt)
  }

  return(out_list)
}

# Example usage
CharmObj_bindingvalues <- build_bindingvalues_list()




restructure_bindingvalues_full <- function(bindingvalues_list) {
  out <- list()
  
  for (rbp_name in names(bindingvalues_list)) {
    df <- bindingvalues_list[[rbp_name]]$bindingvalues
    if (is.null(df) || nrow(df) == 0) next
    
    setDT(df)  # ensure it's a data.table
    
    # Group by Target, dPSI, Metric → store only Pos & Value
    nested_list <- df[, .(data = list(.SD[, .(Pos, Value)])), 
                      by = .(Target, dPSI, Metric)]
    
    # Now structure as: RBP → Target → dPSI → Metric → data.frame
    rbp_nested <- split(nested_list, by = "Target", keep.by = FALSE)
    
    rbp_nested <- lapply(rbp_nested, function(target_dt) {
      target_split <- split(target_dt, by = "dPSI", keep.by = FALSE)
      
      lapply(target_split, function(dpsi_dt) {
        setNames(dpsi_dt$data, dpsi_dt$Metric)
      })
    })
    
    out[[rbp_name]] <- rbp_nested
  }
  
  return(out)
}

# Example usage:
CharmObj_bindingvalues_nested <- restructure_bindingvalues_full(CharmObj_bindingvalues)

CharmObj_bindingvalues <- readRDS("~/Projects/CHARM/data/Binding_both.RDS")
```

# D. RBP Map Basis

```{r}
eCLIPSE_full <- function(bindingvalues_nested, 
                         rnaBP, 
                         target, 
                         dPSI,
                         metric = "FDR", 
                         title = NULL, 
                         plot = TRUE) {
  
  # --- 1. Validate RBP and target
  if (!rnaBP %in% names(bindingvalues_nested)) 
    stop(paste("RBP", rnaBP, "not found."))
  
  if (!target %in% names(bindingvalues_nested[[rnaBP]]))
    stop(paste("Target", target, "not found for", rnaBP))
  
  # --- 2. Fuzzy numeric matching for dPSI
  available_dpsis <- names(bindingvalues_nested[[rnaBP]][[target]])
  available_nums <- suppressWarnings(as.numeric(sub(" .*", "", available_dpsis)))
  dpsi_numeric <- as.numeric(dPSI)
  
  if (is.na(dpsi_numeric))
    stop("dPSI must be numeric or numeric-like, e.g. '0.05'")
  
  diff_vals <- abs(available_nums - dpsi_numeric)
  if (all(is.na(diff_vals)))
    stop(paste("No valid numeric dPSIs found for", rnaBP, "-", target))
  
  best_idx <- which.min(diff_vals)
  chosen_dpsi_name <- available_dpsis[best_idx]
  
  message("→ Using dPSI = ", chosen_dpsi_name, 
          " (closest to requested ", dPSI, ")")
  
  # --- 3. Extract nested data
  this_data <- bindingvalues_nested[[rnaBP]][[target]][[chosen_dpsi_name]]
  
  # --- 4. Extract raw event data
  raw_inc   <- this_data$raw$IncreasedEvents
  raw_dec   <- this_data$raw$DecreasedEvents
  raw_maint <- this_data$raw$MaintainedEvents
  
  # --- 5. Extract summary metrics (new)
  get_value <- function(df) {
    if (is.null(df) || nrow(df) == 0) return(NA)
    if ("Value" %in% names(df)) return(df$Value[1])
    as.numeric(df[1, 1])
  }
  
  inc_target  <- get_value(this_data$raw$IncreasedTargetEvents)
  inc_total   <- get_value(this_data$raw$TotalIncreasedEvents)
  maint_target <- get_value(this_data$raw$MaintainedTargetEvents)
  maint_total  <- get_value(this_data$raw$TotalMaintainedEvents)
  dec_target  <- get_value(this_data$raw$DecreasedTargetEvents)
  dec_total   <- get_value(this_data$raw$TotalDecreasedEvents)
  
  # --- 6. Extract pval/oddsrat data
  pval_inc <- this_data$pval$pvalinc
  pval_dec <- this_data$pval$pvaldec
  odds_inc <- this_data$oddsrat$oddsratinc
  odds_dec <- this_data$oddsrat$oddsratdec
  
  # --- 7. Build data frames for plotting
  dfforvis_raw <- data.frame(
    Pos = raw_inc$Pos,
    IncreasedEvents = raw_inc$Value,
    DecreasedEvents = raw_dec$Value,
    MaintainedEvents = raw_maint$Value
  )
  
  dfforvis_pval <- data.frame(
    pos = pval_inc$Pos,
    pvalinc = pval_inc$Value,
    pvaldec = pval_dec$Value
  )
  
  dfforvis_oddsratio <- data.frame(
    pos = odds_inc$Pos,
    oddsratinc = odds_inc$Value,
    oddsratdec = odds_dec$Value
  )
  
  # --- 8. Plot main binding density
  mapplot <- ggplot(dfforvis_raw, aes(x = Pos)) +
    geom_line(aes(y = IncreasedEvents, color = "IncreasedEvents"), size = 1) +
    geom_line(aes(y = DecreasedEvents, color = "DecreasedEvents"), size = 1) +
    geom_line(aes(y = MaintainedEvents, color = "MaintainedEvents"), size = 1) +
    xlim(0, 1000) +
    labs(title = paste0(rnaBP, " RNA Binding Map - ", title),
         x = "",
         y = " Normalised\nDensity\n") +
    scale_color_manual(values = c(
      IncreasedEvents = "#de425b", 
      DecreasedEvents = "#769fca", 
      MaintainedEvents = "black"
    )) +
    theme_bw() +
    geom_vline(xintercept = c(0,50,250,450,500,550,750,950,1000), linetype = "dashed") +
    theme(
      axis.line = element_line(colour = "black"),
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      panel.background = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      legend.position = "none",
      text = element_text(size = 15)
    )
  
  # --- 9. New label section with summary counts
  labels_plot <- ggplot() +
    annotate("text", x = 1, y = 0,
             label = paste0("Increased: ", inc_target, " (", inc_total, ")\n"),
             color = "#de425b", size = 4, fontface = "bold", hjust = -0.75) +
    annotate("text", x = 1, y = 0,
             label = paste0("Maintained: ", maint_target, " (", maint_total, ")\n"),
             color = "black", size = 4, fontface = "bold", hjust = 0.5) +
    annotate("text", x = 1, y = 0,
             label = paste0("Decreased: ", dec_target, " (", dec_total, ")\n"),
             color = "#769fca", size = 4, fontface = "bold", hjust = 1.75) +
    theme_void() +
    theme(plot.margin = margin(t = 0, r = 0, b = 0, l = 0))
  
  # --- 10. Metric-specific plots
  if (metric == "EffectSize") {
    metricplot <- ggplot(dfforvis_oddsratio, aes(x = pos)) +
      geom_line(aes(y = oddsratinc, color = "Increased"), size = 1) +
      geom_line(aes(y = oddsratdec, color = "Decreased"), size = 1) +
      xlim(0, 1000) +
      labs(x = "Position", y = "Chi-Squared\nStatistic\n") +
      theme_bw() +
      geom_vline(xintercept = c(0,50,250,450,500,550,750,950,1000), linetype = "dashed") +
      scale_color_manual(values = c("Increased" = "#de425b", "Decreased" = "#769fca")) +
      theme(
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "none",
        text = element_text(size = 15)
      ) +
      scale_y_continuous(labels = scales::label_number(accuracy = 0.01))
  } else if (metric == "FDR") {
    metricplot <- ggplot(dfforvis_pval, aes(x = pos)) +
      geom_line(aes(y = pvalinc, color = "Increased"), size = 1) +
      geom_line(aes(y = pvaldec, color = "Decreased"), size = 1) +
      xlim(0, 1000) +
      labs(x = "Position", y = "-log10 (FDR)\n") +
      theme_bw() +
      geom_vline(xintercept = c(0,50,250,450,500,550,750,950,1000), linetype = "dashed") +
      scale_color_manual(values = c("Increased" = "#de425b", "Decreased" = "#769fca")) +
      theme(
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "none",
        text = element_text(size = 15)
      ) +
      scale_y_continuous(labels = scales::label_number(accuracy = 0.01))
  } else {
    stop("metric must be 'FDR' or 'EffectSize'")
  }
  
  # --- 11. Combine plots
  completeplot <- ggarrange(
    mapplot, labels_plot, metricplot,
    ncol = 1, heights = c(0.4, 0.1, 0.4), align = "v"
  )
  
  if (plot) {
    return(completeplot)
  } else {
    return(list(pval_data = dfforvis_pval, oddsratio_data = dfforvis_oddsratio))
  }
}

plot1 <- eCLIPSE_full(
  bindingvalues_nested = CharmObj_bindingvalues,
  rnaBP = "TIA1",
  target = "G3BP1",
  dPSI = "0.05",
  metric = "FDR",
  title = "G3BP1"
)
plot1
```

# E. RBP Similarity

## File Preparation Part 1 

First, I will prepare the files by maintaining only the psi that maximise odds ratio.

```{r fig.height=20, fig.width=20}
Charm.object.binding.ES <- qs::qread("../data/QS_Files/Binding_K562.qs")

filter_binding_dpsi <- function(binding_obj) {

  keep_pattern <- "(maximised oddsratinc|maximised oddsratdec)"

  out <- binding_obj

  for (rbp in names(out)) {
    for (tgt in names(out[[rbp]])) {

      dpsi_names <- names(out[[rbp]][[tgt]])
      if (is.null(dpsi_names)) next

      keep <- dpsi_names[
        grepl(keep_pattern, dpsi_names, ignore.case = TRUE) |
        grepl("(^|[^0-9])0\\.05([^0-9]|$)", dpsi_names) |
        grepl("(^|[^0-9])0\\.10([^0-9]|$)", dpsi_names)
      ]

      if (length(keep) == 0) {
        out[[rbp]][[tgt]] <- NULL
      } else {
        out[[rbp]][[tgt]] <- out[[rbp]][[tgt]][keep]
      }
    }

    if (length(out[[rbp]]) == 0) {
      out[[rbp]] <- NULL
    }
  }

  out
}
filter_binding_keep_oddsrat <- function(binding_obj) {

  out <- binding_obj

  for (rbp in names(out)) {
    for (tgt in names(out[[rbp]])) {
      for (dpsi in names(out[[rbp]][[tgt]])) {

        dpsi_block <- out[[rbp]][[tgt]][[dpsi]]

        # If oddsrat does not exist, drop this dPSI
        if (!is.list(dpsi_block) || !"oddsrat" %in% names(dpsi_block)) {
          out[[rbp]][[tgt]][[dpsi]] <- NULL
          next
        }

        # Safely extract raw event counts (may be NULL)
        inc_events <- NULL
        dec_events <- NULL

        if ("raw" %in% names(dpsi_block)) {
          raw_block <- dpsi_block[["raw"]]

          if (is.list(raw_block)) {
            inc_events <- raw_block[["IncreasedTargetEvents"]]
            dec_events <- raw_block[["DecreasedTargetEvents"]]
          }
        }

        # Replace dPSI block with trimmed content
        out[[rbp]][[tgt]][[dpsi]] <- list(
          oddsrat = dpsi_block[["oddsrat"]],
          IncreasedTargetEvents = inc_events,
          DecreasedTargetEvents = dec_events
        )
      }

      # Drop empty targets
      if (length(out[[rbp]][[tgt]]) == 0) {
        out[[rbp]][[tgt]] <- NULL
      }
    }

    # Drop empty RBPs
    if (length(out[[rbp]]) == 0) {
      out[[rbp]] <- NULL
    }
  }

  out
}

Charm.binding.ES.filtered <- filter_binding_dpsi(Charm.object.binding.ES)
Charm.binding.ES.filtered <- filter_binding_keep_oddsrat(Charm.binding.ES.filtered)
#rm(Charm.object.binding.ES)
```

Great, now I will calculate how much signal each pair of RBP and Target has, at selected dPSIs.

```{r}
summarise_oddsrat_abs_sum <- function(binding_obj) {

  out <- list()

  for (rbp in names(binding_obj)) {
    out[[rbp]] <- list()

    for (tgt in names(binding_obj[[rbp]])) {
      out[[rbp]][[tgt]] <- list()

      for (dpsi in names(binding_obj[[rbp]][[tgt]])) {

        dpsi_block <- binding_obj[[rbp]][[tgt]][[dpsi]]
        if (!is.list(dpsi_block)) next

        oddsrat_block <- dpsi_block[["oddsrat"]]
        if (is.null(oddsrat_block)) next

        # --- extract event counts safely ---
        inc_tbl <- dpsi_block[["IncreasedTargetEvents"]]
        dec_tbl <- dpsi_block[["DecreasedTargetEvents"]]

        n_inc <- if (is.data.frame(inc_tbl) && "Value" %in% colnames(inc_tbl))
          inc_tbl$Value[1] else 0

        n_dec <- if (is.data.frame(dec_tbl) && "Value" %in% colnames(dec_tbl))
          dec_tbl$Value[1] else 0

        dpsi_out <- list()

        # -------------------------
        # Decreased
        # -------------------------
        df_dec <- oddsrat_block[["oddsratdec"]]
        if (is.data.frame(df_dec) && n_dec > 0) {
          dpsi_out[["oddsratdec"]] <-
            sum(abs(df_dec$Value), na.rm = TRUE) * (n_dec^(1/2))
        }

        # -------------------------
        # Increased
        # -------------------------
        df_inc <- oddsrat_block[["oddsratinc"]]
        if (is.data.frame(df_inc) && n_inc > 0) {
          dpsi_out[["oddsratinc"]] <-
            sum(abs(df_inc$Value), na.rm = TRUE) * (n_inc^(1/2))
        }

        if (length(dpsi_out) > 0) {
          out[[rbp]][[tgt]][[dpsi]] <- dpsi_out
        }
      }

      if (length(out[[rbp]][[tgt]]) == 0)
        out[[rbp]][[tgt]] <- NULL
    }

    if (length(out[[rbp]]) == 0)
      out[[rbp]] <- NULL
  }

  out
}


Charm.binding.oddsrat.sum <- summarise_oddsrat_abs_sum(Charm.binding.ES.filtered)

extract_oddsrat_values <- function(oddsrat_obj) {

  inc_vals <- numeric()
  dec_vals <- numeric()

  for (rbp in names(oddsrat_obj)) {
    for (tgt in names(oddsrat_obj[[rbp]])) {
      for (dpsi in names(oddsrat_obj[[rbp]][[tgt]])) {

        block <- oddsrat_obj[[rbp]][[tgt]][[dpsi]]

        if (!is.null(block$oddsratinc)) {
          inc_vals <- c(inc_vals, block$oddsratinc)
        }
        if (!is.null(block$oddsratdec)) {
          dec_vals <- c(dec_vals, block$oddsratdec)
        }
      }
    }
  }

  list(
    oddsratinc = inc_vals,
    oddsratdec = dec_vals
  )
}

plot_oddsrat_densities <- function(oddsrat_obj,
                                   inc_color = "red",
                                   dec_color = "blue",
                                   bw = "nrd0") {

  vals <- extract_oddsrat_values(oddsrat_obj)

  if (length(vals$oddsratinc) < 2 || length(vals$oddsratdec) < 2) {
    stop("Not enough values to compute densities.")
  }

  dens_inc <- density(log10(vals$oddsratinc), bw = bw)
  dens_dec <- density(log10(vals$oddsratdec), bw = bw)

  # ---- Increased ----
  plot(
    dens_inc,
    col = inc_color,
    lwd = 3,
    main = "Distribution of |Odds Ratio| – Increased Events",
    xlab = "log10|Odds Ratio| Integral",
    ylab = "Density"
  )+abline(v=4.5, col="black")

  # ---- Decreased ----
  plot(
    dens_dec,
    col = dec_color,
    lwd = 3,
    main = "Distribution of |Odds Ratio| – Decreased Events",
    xlab = "log10|Odds Ratio| Integral",
    ylab = "Density"
  )
}

plot_oddsrat_densities(Charm.binding.oddsrat.sum)


```

 Counting the before and after:
 
```{r}
filter_and_select_top_oddsrat <- function(oddsrat_obj,
                                          log10_threshold = 6) {

  filtered <- list()
  n_inc <- 0L
  n_dec <- 0L

  for (rbp in names(oddsrat_obj)) {
    for (tgt in names(oddsrat_obj[[rbp]])) {

      best_inc <- NULL
      best_inc_dpsi <- NULL

      best_dec <- NULL
      best_dec_dpsi <- NULL

      for (dpsi in names(oddsrat_obj[[rbp]][[tgt]])) {

        block <- oddsrat_obj[[rbp]][[tgt]][[dpsi]]

        ## oddsratinc
        if (!is.null(block$oddsratinc) &&
            is.finite(block$oddsratinc) &&
            log10(block$oddsratinc) > log10_threshold) {

          if (is.null(best_inc) || block$oddsratinc > best_inc) {
            best_inc <- block$oddsratinc
            best_inc_dpsi <- dpsi
          }
        }

        ## oddsratdec
        if (!is.null(block$oddsratdec) &&
            is.finite(block$oddsratdec) &&
            log10(block$oddsratdec) > log10_threshold) {

          if (is.null(best_dec) || block$oddsratdec > best_dec) {
            best_dec <- block$oddsratdec
            best_dec_dpsi <- dpsi
          }
        }
      }

      ## Store results (at most 1 inc + 1 dec)
      if (!is.null(best_inc) || !is.null(best_dec)) {

        filtered[[rbp]][[tgt]] <- list()

        if (!is.null(best_inc)) {
          filtered[[rbp]][[tgt]][[best_inc_dpsi]]$oddsratinc <- best_inc
          n_inc <- n_inc + 1L
        }

        if (!is.null(best_dec)) {
          filtered[[rbp]][[tgt]][[best_dec_dpsi]]$oddsratdec <- best_dec
          n_dec <- n_dec + 1L
        }
      }
    }
  }

  list(
    counts = c(
      oddsratinc = n_inc,
      oddsratdec = n_dec
    ),
    filtered_object = filtered
  )
}

res <- filter_and_select_top_oddsrat(
  Charm.binding.oddsrat.sum,
  log10_threshold = -20
)

res$counts
#res$filtered_object



```

```{r}
filtered_object <- res[["filtered_object"]]

all_targets <- unlist(
  lapply(filtered_object, names),
  use.names = FALSE
)

unique_targets <- (unique(all_targets))

length(unique(all_targets))

filtered_object <- res$filtered_object

filtered_object <- lapply(
  filtered_object,
  function(rbp_block) {
    rbp_block[names(rbp_block) %in% unique_targets]
  }
)


```

```{r}
filtered_object <- res[["filtered_object"]]
flatten_oddsrat <- function(filtered_object,
                            direction = c("oddsratinc", "oddsratdec")) {

  direction <- match.arg(direction)
  out <- list()
  k <- 1L

  for (rbp in names(filtered_object)) {
    for (tgt in names(filtered_object[[rbp]])) {
      for (dpsi in names(filtered_object[[rbp]][[tgt]])) {

        vec <- filtered_object[[rbp]][[tgt]][[dpsi]][[direction]]

        if (is.null(vec) || length(vec) == 0) next
        if (!is.numeric(vec)) next

        out[[k]] <- data.table(
          RBP    = rbp,
          Target = tgt,
          dPSI   = dpsi,
          Value  = vec
        )
        k <- k + 1L
      }
    }
  }

  rbindlist(out, use.names = TRUE)
}

oddsratinc_df <- flatten_oddsrat(
  filtered_object,
  direction = "oddsratinc"
)

oddsratdec_df <- flatten_oddsrat(
  filtered_object,
  direction = "oddsratdec"
)

```


```{r fig.height=20, fig.width=20}
library(data.table)
library(pheatmap)

setnames(
  oddsratinc_df,
  old = "Value",
  new = "oddsratinc"
)

setnames(
  oddsratdec_df,
  old = "Value",
  new = "oddsratdec"
)

rbp_target_heatmap_raw <- function(
  df,
  value_col,
  agg_fun = c("mean", "median"),
  clustering_method = "ward.D2",
  clustering_distance = "euclidean",
  show_rownames = TRUE,
  show_colnames = TRUE,
  main = "RBP × Target heatmap (raw values)"
) {

  agg_fun <- match.arg(agg_fun)
  dt <- as.data.table(df)

  stopifnot(all(c("RBP", "Target", value_col) %in% colnames(dt)))

  # -------------------------------------------------
  # Aggregate ONLY if needed (no transformation)
  # -------------------------------------------------
  agg <- dt[
    ,
    .(
      value = switch(
        agg_fun,
        mean   = mean(get(value_col), na.rm = TRUE),
        median = median(get(value_col), na.rm = TRUE)
      )
    ),
    by = .(RBP, Target)
  ]

  # -------------------------------------------------
  # Build matrix
  # -------------------------------------------------
  rbps    <- sort(unique(agg$RBP))
  targets <- sort(unique(agg$Target))

  S <- matrix(
    NA_real_,
    nrow = length(rbps),
    ncol = length(targets),
    dimnames = list(rbps, targets)
  )

  i <- match(agg$RBP, rbps)
  j <- match(agg$Target, targets)

  S[cbind(i, j)] <- agg$value

  # -------------------------------------------------
  # Plot
  # -------------------------------------------------
  pheatmap(
    log10(S),
    clustering_method = clustering_method,
    clustering_distance_rows = clustering_distance,
    clustering_distance_cols = clustering_distance,
    color = colorRampPalette(c("navy", "white", "firebrick"))(100),
    show_rownames = show_rownames,
    show_colnames = show_colnames,
    border_color = NA,
    main = main
  )

  # S = your matrix/dataframe (rows = RBPs, cols = targets)

max_values <- apply(S, 2, max, na.rm = TRUE)       # maximum per column
max_rows   <- apply(S, 2, function(col) {
  rownames(S)[which.max(col)]
})

# Combine into a dataframe
max_per_column <- data.frame(
  Target = colnames(S),
  MaxValue = max_values,
  MaxRBP = max_rows
)

print(max_per_column)

  
  invisible(S)
}

#oddsratinc_df_alter <- 

rbp_target_heatmap_raw(
  oddsratinc_df,
  value_col = "oddsratinc",
  main = "RBP × Target – Oddsratinc (raw)"
)

rbp_target_heatmap_raw(
  oddsratdec_df,
  value_col = "oddsratdec",
  main = "RBP × Target – Oddsratdec (raw)"
)

```

```{r fig.height=20, fig.width=20}
library(data.table)
library(pheatmap)

rbp_target_heatmap_raw <- function(
  df,
  value_col,
  agg_fun = c("mean", "median"),
  clustering_method = "ward.D2",
  clustering_distance = "euclidean",
  show_rownames = TRUE,
  show_colnames = TRUE,
  main = "RBP × Target heatmap (raw values)"
) {

  agg_fun <- match.arg(agg_fun)
  dt <- as.data.table(df)

  stopifnot(all(c("RBP", "Target", value_col) %in% colnames(dt)))

  # -------------------------------------------------
  # Aggregate ONLY if needed (no transformation)
  # -------------------------------------------------
  agg <- dt[
    ,
    .(
      value = switch(
        agg_fun,
        mean   = mean(get(value_col), na.rm = TRUE),
        median = median(get(value_col), na.rm = TRUE)
      )
    ),
    by = .(RBP, Target)
  ]

  # -------------------------------------------------
  # Build matrix
  # -------------------------------------------------
  rbps    <- sort(unique(agg$RBP))
  targets <- sort(unique(agg$Target))

  S <- matrix(
    NA_real_,
    nrow = length(rbps),
    ncol = length(targets),
    dimnames = list(rbps, targets)
  )

  i <- match(agg$RBP, rbps)
  j <- match(agg$Target, targets)

  S[cbind(i, j)] <- agg$value

  # -------------------------------------------------
  # Plot
  # -------------------------------------------------
S_log <- log10(S + 1)  # add 1 to avoid log(0)
S_z <- scale(S_log, center = TRUE, scale = TRUE)

# Keep as a matrix for pheatmap
S_z <- as.matrix(S_z)

  
pheatmap(
  S_z,
  cluster_rows = TRUE,   # cluster RBPs
  cluster_cols = TRUE,  # keep targets in the original order
  clustering_method = "ward.D2",
  color = colorRampPalette(c("navy", "white", "firebrick"))(100),
  show_rownames = TRUE,
  show_colnames = TRUE,
  border_color = NA,
  main = main
)

  # S = your matrix/dataframe (rows = RBPs, cols = targets)

max_values <- apply(S, 2, max, na.rm = TRUE)       # maximum per column
max_rows   <- apply(S, 2, function(col) {
  rownames(S)[which.max(col)]
})

# Combine into a dataframe
max_per_column <- data.frame(
  Target = colnames(S),
  MaxValue = max_values,
  MaxRBP = max_rows
)

print(max_per_column)

  
  invisible(S)
}

#oddsratinc_df_alter <- 

rbp_target_heatmap_raw(
  oddsratinc_df,
  value_col = "oddsratinc",
  main = "RBP × Target – log10Oddsratinc (Target zScore)"
)

rbp_target_heatmap_raw(
  oddsratdec_df,
  value_col = "oddsratdec",
  main = "RBP × Target – log10Oddsratdec (Target zScore)"
)

```

```{r fig.height=20, fig.width=20}
rbp_target_pca <- function(
  df,
  value_col,
  agg_fun = c("mean", "median"),
  log_transform = TRUE,
  zscore_columns = TRUE,
  min_non_na = 2,
  main = "PCA – RBPs (Targets as genes)"
) {

  agg_fun <- match.arg(agg_fun)
  dt <- as.data.table(df)

  stopifnot(all(c("RBP", "Target", value_col) %in% colnames(dt)))

  # -------------------------------------------------
  # Aggregate values per RBP–Target
  # -------------------------------------------------
  agg <- dt[
    ,
    .(
      value = switch(
        agg_fun,
        mean   = mean(get(value_col), na.rm = TRUE),
        median = median(get(value_col), na.rm = TRUE)
      )
    ),
    by = .(RBP, Target)
  ]

  # -------------------------------------------------
  # Build matrix: rows = RBPs, cols = Targets
  # -------------------------------------------------
  S <- dcast(agg, RBP ~ Target, value.var = "value")

  rbp_names <- S$RBP
  S$RBP <- NULL
  S <- as.matrix(S)
  rownames(S) <- rbp_names

  # Replace NAs with 0 (or choose another strategy if desired)
  S[is.na(S)] <- 0

  # -------------------------------------------------
  # Transformations
  # -------------------------------------------------
  if (log_transform) {
    S <- log10(S + 1)
  }

  if (zscore_columns) {
    S <- scale(S, center = TRUE, scale = TRUE)
  }

  S <- as.matrix(S)

  # -------------------------------------------------
  # Remove zero-variance Targets
  # -------------------------------------------------
  keep_cols <- apply(S, 2, function(x)
    sum(is.finite(x)) >= min_non_na && sd(x, na.rm = TRUE) > 0
  )

  S <- S[, keep_cols, drop = FALSE]

  # -------------------------------------------------
  # PCA (RBPs are observations)
  # -------------------------------------------------
  pca_res <- prcomp(S, center = FALSE, scale. = FALSE)

  var_explained <- (pca_res$sdev^2) / sum(pca_res$sdev^2)

  pca_df <- data.frame(
    RBP = rownames(S),
    PC1 = pca_res$x[, 1],
    PC2 = pca_res$x[, 2]
  )

  # -------------------------------------------------
  # Plot
  # -------------------------------------------------
  p <- ggplot(pca_df, aes(PC1, PC2)) +
    geom_point(size = 4, alpha = 0.8) +
    geom_text_repel(aes(label = RBP), size = 3) +
    theme_minimal(base_size = 14) +
    labs(
      title = main,
      x = paste0("PC1 (", round(var_explained[1] * 100, 1), "%)"),
      y = paste0("PC2 (", round(var_explained[2] * 100, 1), "%)")
    )

  print(p)

  invisible(list(
    matrix = S,
    pca = pca_res,
    variance = var_explained,
    plot = p
  ))
}

rbp_target_pca(
  oddsratinc_df,
  value_col = "oddsratinc",
  main = "PCA – RBPs (Oddsratinc)"
)

rbp_target_pca(
  oddsratdec_df,
  value_col = "oddsratdec",
  main = "PCA – RBPs (Oddsratdec)"
)

```


## Filtering
 
```{r}
filter_original_binding_by_oddsrat <- function(original_binding,
                                               filtered_object) {

  out <- list()

  for (rbp in names(filtered_object)) {
    if (!rbp %in% names(original_binding)) next

    out[[rbp]] <- list()

    for (tgt in names(filtered_object[[rbp]])) {
      if (!tgt %in% names(original_binding[[rbp]])) next

      out[[rbp]][[tgt]] <- list()

      for (dpsi in names(filtered_object[[rbp]][[tgt]])) {
        if (!dpsi %in% names(original_binding[[rbp]][[tgt]])) next

        oddsrat_block <- original_binding[[rbp]][[tgt]][[dpsi]][["oddsrat"]]
        if (is.null(oddsrat_block)) next

        keep_dirs <- filtered_object[[rbp]][[tgt]][[dpsi]]
        new_oddsrat <- list()

        for (dir in names(keep_dirs)) {
          if (dir %in% names(oddsrat_block)) {
            new_oddsrat[[dir]] <- oddsrat_block[[dir]]
          }
        }

        if (length(new_oddsrat) > 0) {
          out[[rbp]][[tgt]][[dpsi]] <- list(
            oddsrat = new_oddsrat
          )
        }
      }

      if (length(out[[rbp]][[tgt]]) == 0) {
        out[[rbp]][[tgt]] <- NULL
      }
    }

    if (length(out[[rbp]]) == 0) {
      out[[rbp]] <- NULL
    }
  }

  out
}

Charm.object.binding.filtered <- filter_original_binding_by_oddsrat(
  original_binding = Charm.object.binding.ES,
  filtered_object  = res$filtered_object
)


filter_original_binding_by_oddsrat <- function(original_binding,
                                               filtered_object) {

  out <- list()

  for (rbp in names(filtered_object)) {
    if (!rbp %in% names(original_binding)) next

    out[[rbp]] <- list()

    for (tgt in names(filtered_object[[rbp]])) {
      if (!tgt %in% names(original_binding[[rbp]])) next

      out[[rbp]][[tgt]] <- list()

      for (dpsi in names(filtered_object[[rbp]][[tgt]])) {
        if (!dpsi %in% names(original_binding[[rbp]][[tgt]])) next

        block <- original_binding[[rbp]][[tgt]][[dpsi]]

        oddsrat_block <- block[["oddsrat"]]
        raw_block     <- block[["raw"]]

        if (is.null(oddsrat_block) || is.null(raw_block)) next

        # -------------------------------------------------
        # Extract event scalars correctly
        # -------------------------------------------------
        inc_events <- raw_block[["IncreasedTargetEvents"]][["Value"]]
        dec_events <- raw_block[["DecreasedTargetEvents"]][["Value"]]

        keep_dirs <- filtered_object[[rbp]][[tgt]][[dpsi]]
        new_oddsrat <- list()

        for (dir in names(keep_dirs)) {

          if (!dir %in% names(oddsrat_block)) next

          df <- oddsrat_block[[dir]]
          if (!is.data.frame(df) || !"Value" %in% colnames(df)) next

          df_new <- df

          # -------------------------------------------------
          # Multiply correctly
          # -------------------------------------------------
          if (dir == "oddsratinc" && !is.null(inc_events)) {
            df_new$Value <- df_new$Value * inc_events
          }

          if (dir == "oddsratdec" && !is.null(dec_events)) {
            df_new$Value <- df_new$Value * dec_events
          }

          new_oddsrat[[dir]] <- df_new
        }

        if (length(new_oddsrat) > 0) {
          out[[rbp]][[tgt]][[dpsi]] <- list(
            oddsrat = new_oddsrat
          )
        }
      }

      if (length(out[[rbp]][[tgt]]) == 0) {
        out[[rbp]][[tgt]] <- NULL
      }
    }

    if (length(out[[rbp]]) == 0) {
      out[[rbp]] <- NULL
    }
  }

  out
}

Charm.object.binding.filtered <- filter_original_binding_by_oddsrat(
  original_binding = Charm.object.binding.ES,
  filtered_object  = res$filtered_object
)


saveRDS(Charm.object.binding.filtered, "../intermediary_files/Charm.binding.filtered.both_new_full_K562.Rds")
```
 
 
## Creating python format

```{r}
oddsrat_list_to_dataframe <- function(filtered_obj,
                                      n_pos = 1000,
                                      prefix = "pos_") {

  rows <- vector("list", 50000)
  idx <- 1L

  for (rbp in names(filtered_obj)) {
    for (tgt in names(filtered_obj[[rbp]])) {
      for (dpsi in names(filtered_obj[[rbp]][[tgt]])) {

        block <- filtered_obj[[rbp]][[tgt]][[dpsi]]

        ## Safety
        if (is.null(block$oddsrat))
          next

        ## ---------- oddsratinc ----------
        if (!is.null(block$oddsrat$oddsratinc)) {

          v <- block$oddsrat$oddsratinc$Value

          if (length(v) != n_pos)
            stop(paste("Unexpected length for oddsratinc:",
                       rbp, tgt, dpsi))

          rows[[idx]] <- c(
            RBP = rbp,
            Target = tgt,
            dPSI = dpsi,
            Direction = "oddsratinc",
            setNames(as.list(v), paste0(prefix, seq_len(n_pos)))
          )

          idx <- idx + 1L
        }

        ## ---------- oddsratdec ----------
        if (!is.null(block$oddsrat$oddsratdec)) {

          v <- block$oddsrat$oddsratdec$Value

          if (length(v) != n_pos)
            stop(paste("Unexpected length for oddsratdec:",
                       rbp, tgt, dpsi))

          rows[[idx]] <- c(
            RBP = rbp,
            Target = tgt,
            dPSI = dpsi,
            Direction = "oddsratdec",
            setNames(as.list(v), paste0(prefix, seq_len(n_pos)))
          )

          idx <- idx + 1L
        }
      }
    }
  }

  if (idx == 1L)
    stop("No oddsrat entries found — check structure again")

  df <- as.data.frame(
    do.call(rbind, rows[seq_len(idx - 1)]),
    stringsAsFactors = FALSE
  )

  ## Convert positional columns back to numeric
  pos_cols <- grep(paste0("^", prefix), colnames(df))
  df[, pos_cols] <- lapply(df[, pos_cols], as.numeric)

  df
}
oddsrat_df <- oddsrat_list_to_dataframe(
  Charm.object.binding.filtered,
  n_pos = 1000   # or 500 for IR
)

data.table::fwrite(
  oddsrat_df,
  "../intermediary_files/Charm.binding.filtered.both_new_full_K562.csv",
  sep = "\t"
)

```


# F. Alternative Method (shRNA)

## Increased

```{r}

dt <- fread(
  "/mnt/scratch/home/alexandre/Projects/CHARM/intermediary_files/FullSimilarity_oddsratinc.csv"
)

dt <- unique(dt)

# Force RBP1 < RBP2 ordering
dt[, `:=`(
  RBP_low  = pmin(RBP1, RBP2),
  RBP_high = pmax(RBP1, RBP2)
)]

dt[, Target_low  := ifelse(RBP1 < RBP2, Target1, Target2)]
dt[, Target_high := ifelse(RBP1 < RBP2, Target2, Target1)]

result <- dt[
  ,
  .(
    Cosine = mean(Cosine, na.rm = TRUE),
    n_rows = uniqueN(c(Target1, Target2))  # counts unique targets across both RBPs
  ),
  by = .(RBP1 = RBP_low, RBP2 = RBP_high)
]


```
### Heatmap

```{r fig.height=20, fig.width=20}
dt <- copy(result)

rbps <- sort(unique(c(dt$RBP1, dt$RBP2)))
n <- length(rbps)

rbp_index <- setNames(seq_len(n), rbps)

S <- matrix(NA_real_, n, n,
            dimnames = list(rbps, rbps))

i <- rbp_index[dt$RBP1]
j <- rbp_index[dt$RBP2]

S[cbind(i, j)] <- dt$Cosine
S[cbind(j, i)] <- dt$Cosine
#diag(S) <- 1

S[is.na(S)] <- 0

D <- 1 - S

pheatmap(
  S,
  clustering_method = "ward.D2",
  color = colorRampPalette(c("navy", "white", "firebrick"))(100),
  show_rownames = TRUE,
  show_colnames = TRUE,
  border_color = NA,
  main = "shRNA Similarity – Increased Events"
)



```

### MDS

```{r fig.height=20, fig.width=20}

pair_df <- result

rbp_stats <- rbind(
  pair_df[, .(RBP = RBP1, n_pairs = n_rows)],
  pair_df[, .(RBP = RBP2, n_pairs = n_rows)]
)[
  ,
  .(n_pairs = sum(n_pairs)),
  by = RBP
]

mds <- cmdscale(D, k = 2, eig = TRUE)

mds_df <- data.frame(
  RBP  = rownames(D),
  MDS1 = mds$points[, 1],
  MDS2 = mds$points[, 2],
  stringsAsFactors = FALSE
)

mds_df <- merge(
  mds_df,
  rbp_stats,
  by = "RBP",
  all.x = TRUE
)

mds_df$n_pairs[is.na(mds_df$n_pairs)] <- 0

ggplot(mds_df, aes(MDS1, MDS2)) +
  geom_point(aes(size = n_pairs), alpha = 0.75) +
  geom_text_repel(
    aes(label = RBP),
    size = 3,
    max.overlaps = 30
  ) +
  scale_size_continuous(range = c(2, 10)) +
  theme_minimal(base_size = 14) +
  labs(
    title = "shRNA Similarity – Increased Events",
    size = "Shared eCLIP\ncomparisons"
  )
```

### TSne

```{r fig.height=20, fig.width=20}


# -----------------------------------------
# Choose valid perplexity
# -----------------------------------------
n <- nrow(D)
perp <- min(30, floor((n - 1) / 3))

# -----------------------------------------
# Run t-SNE using distance matrix
# -----------------------------------------
tsne <- Rtsne(
  as.dist(D),        # use distance matrix
  is_distance = TRUE,
  perplexity = perp,
  theta = 0.5,       # speed/accuracy tradeoff
  dims = 2,
  verbose = TRUE
)

tsne_df <- data.frame(
  RBP  = rownames(D),
  tSNE1 = tsne$Y[, 1],
  tSNE2 = tsne$Y[, 2],
  stringsAsFactors = FALSE
)

# -----------------------------------------
# Merge with n_pairs (same as before)
# -----------------------------------------
tsne_df <- merge(
  tsne_df,
  rbp_stats,
  by = "RBP",
  all.x = TRUE
)

tsne_df$n_pairs[is.na(tsne_df$n_pairs)] <- 0

# -----------------------------------------
# Plot
# -----------------------------------------
ggplot(tsne_df, aes(tSNE1, tSNE2)) +
  geom_point(aes(size = n_pairs), alpha = 0.75) +
  geom_text_repel(
    aes(label = RBP),
    size = 3,
    max.overlaps = 30
  ) +
  scale_size_continuous(range = c(2, 10)) +
  theme_minimal(base_size = 14) +
  labs(
    title = "t-SNE – shRNA Similarity (Increased Events)",
    size = "Shared eCLIP\ncomparisons"
  )
```


### NRows

```{r fig.height=20, fig.width=20}
# result has: RBP1, RBP2, Wasserstein, n_rows

pair_df <- copy(result)

rbps <- sort(unique(c(pair_df$RBP1, pair_df$RBP2)))
n <- length(rbps)

rbp_index <- setNames(seq_len(n), rbps)

Nmat <- matrix(
  0,
  nrow = n,
  ncol = n,
  dimnames = list(rbps, rbps)
)

i <- rbp_index[pair_df$RBP1]
j <- rbp_index[pair_df$RBP2]

Nmat[cbind(i, j)] <- pair_df$n_rows
Nmat[cbind(j, i)] <- pair_df$n_rows   # symmetric

pheatmap(
  Nmat,
  clustering_method = "ward.D2",
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  color = colorRampPalette(c("navy", "white", "firebrick"))(100),
  show_rownames = TRUE,
  show_colnames = TRUE,
  border_color = NA,
  main = "shRNA #Pairs – Increased Events"
)

```


## Decreased

```{r}

dt <- fread(
  "/mnt/scratch/home/alexandre/Projects/CHARM/intermediary_files/FullSimilarity_oddsratdec.csv"
)

dt <- unique(dt)

# Force RBP1 < RBP2 ordering
dt[, `:=`(
  RBP_low  = pmin(RBP1, RBP2),
  RBP_high = pmax(RBP1, RBP2)
)]

dt[, Target_low  := ifelse(RBP1 < RBP2, Target1, Target2)]
dt[, Target_high := ifelse(RBP1 < RBP2, Target2, Target1)]

result <- dt[
  ,
  .(
    Cosine = mean(Cosine, na.rm = TRUE),
    n_rows = uniqueN(c(Target1, Target2))  # counts unique targets across both RBPs
  ),
  by = .(RBP1 = RBP_low, RBP2 = RBP_high)
]


```
### Heatmap

```{r fig.height=20, fig.width=20}
dt <- copy(result)

rbps <- sort(unique(c(dt$RBP1, dt$RBP2)))
n <- length(rbps)

rbp_index <- setNames(seq_len(n), rbps)

S <- matrix(NA_real_, n, n,
            dimnames = list(rbps, rbps))

i <- rbp_index[dt$RBP1]
j <- rbp_index[dt$RBP2]

S[cbind(i, j)] <- dt$Cosine
S[cbind(j, i)] <- dt$Cosine
#diag(S) <- 1

S[is.na(S)] <- 0

D <- 1 - S

pheatmap(
  S,
  clustering_method = "ward.D2",
  color = colorRampPalette(c("navy", "white", "firebrick"))(100),
  show_rownames = TRUE,
  show_colnames = TRUE,
  border_color = NA,
  main = "shRNA Similarity – Decreased Events"
)



```

### MDS

```{r fig.height=20, fig.width=20}

pair_df <- result

rbp_stats <- rbind(
  pair_df[, .(RBP = RBP1, n_pairs = n_rows)],
  pair_df[, .(RBP = RBP2, n_pairs = n_rows)]
)[
  ,
  .(n_pairs = sum(n_pairs)),
  by = RBP
]

mds <- cmdscale(D, k = 2, eig = TRUE)

mds_df <- data.frame(
  RBP  = rownames(D),
  MDS1 = mds$points[, 1],
  MDS2 = mds$points[, 2],
  stringsAsFactors = FALSE
)

mds_df <- merge(
  mds_df,
  rbp_stats,
  by = "RBP",
  all.x = TRUE
)

mds_df$n_pairs[is.na(mds_df$n_pairs)] <- 0

ggplot(mds_df, aes(MDS1, MDS2)) +
  geom_point(aes(size = n_pairs), alpha = 0.75) +
  geom_text_repel(
    aes(label = RBP),
    size = 3,
    max.overlaps = 30
  ) +
  scale_size_continuous(range = c(2, 10)) +
  theme_minimal(base_size = 14) +
  labs(
    title = "shRNA Similarity – Decreased Events",
    size = "Shared eCLIP\ncomparisons"
  )
```


### TSne

```{r fig.height=20, fig.width=20}


# -----------------------------------------
# Choose valid perplexity
# -----------------------------------------
n <- nrow(D)
perp <- min(30, floor((n - 1) / 3))

# -----------------------------------------
# Run t-SNE using distance matrix
# -----------------------------------------
tsne <- Rtsne(
  as.dist(D),        # use distance matrix
  is_distance = TRUE,
  perplexity = perp,
  theta = 0.5,       # speed/accuracy tradeoff
  dims = 2,
  verbose = TRUE
)

tsne_df <- data.frame(
  RBP  = rownames(D),
  tSNE1 = tsne$Y[, 1],
  tSNE2 = tsne$Y[, 2],
  stringsAsFactors = FALSE
)

# -----------------------------------------
# Merge with n_pairs (same as before)
# -----------------------------------------
tsne_df <- merge(
  tsne_df,
  rbp_stats,
  by = "RBP",
  all.x = TRUE
)

tsne_df$n_pairs[is.na(tsne_df$n_pairs)] <- 0

# -----------------------------------------
# Plot
# -----------------------------------------
ggplot(tsne_df, aes(tSNE1, tSNE2)) +
  geom_point(aes(size = n_pairs), alpha = 0.75) +
  geom_text_repel(
    aes(label = RBP),
    size = 3,
    max.overlaps = 30
  ) +
  scale_size_continuous(range = c(2, 10)) +
  theme_minimal(base_size = 14) +
  labs(
    title = "t-SNE – shRNA Similarity (Decreased Events)",
    size = "Shared eCLIP\ncomparisons"
  )
```

### NRows

```{r fig.height=20, fig.width=20}
# result has: RBP1, RBP2, Wasserstein, n_rows

pair_df <- copy(result)

rbps <- sort(unique(c(pair_df$RBP1, pair_df$RBP2)))
n <- length(rbps)

rbp_index <- setNames(seq_len(n), rbps)

Nmat <- matrix(
  0,
  nrow = n,
  ncol = n,
  dimnames = list(rbps, rbps)
)

i <- rbp_index[pair_df$RBP1]
j <- rbp_index[pair_df$RBP2]

Nmat[cbind(i, j)] <- pair_df$n_rows
Nmat[cbind(j, i)] <- pair_df$n_rows   # symmetric

pheatmap(
  Nmat,
  clustering_method = "ward.D2",
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  color = colorRampPalette(c("navy", "white", "firebrick"))(100),
  show_rownames = TRUE,
  show_colnames = TRUE,
  border_color = NA,
  main = "shRNA #Pairs – Decreased Events"
)

```

## Both

```{r}
dt1 <- fread(
  "/mnt/scratch/home/alexandre/Projects/CHARM/intermediary_files/FullSimilarity_oddsratinc.csv"
)

dt2 <- fread(
  "/mnt/scratch/home/alexandre/Projects/CHARM/intermediary_files/FullSimilarity_oddsratdec.csv"
)

dt <- rbind(dt1, dt2)
#dt <- unique(dt)

# Force RBP1 < RBP2 ordering
dt[, `:=`(
  RBP_low  = pmin(RBP1, RBP2),
  RBP_high = pmax(RBP1, RBP2)
)]

dt[, Target_low  := ifelse(RBP1 < RBP2, Target1, Target2)]
dt[, Target_high := ifelse(RBP1 < RBP2, Target2, Target1)]

result <- dt[
  ,
  .(
    Cosine = mean(Cosine, na.rm = TRUE),
    n_rows = uniqueN(c(Target1, Target2))  # counts unique targets across both RBPs
  ),
  by = .(RBP1 = RBP_low, RBP2 = RBP_high)
]


```
### Heatmap

```{r fig.height=20, fig.width=20}
dt <- copy(result)

rbps <- sort(unique(c(dt$RBP1, dt$RBP2)))
n <- length(rbps)

rbp_index <- setNames(seq_len(n), rbps)

S <- matrix(NA_real_, n, n,
            dimnames = list(rbps, rbps))

i <- rbp_index[dt$RBP1]
j <- rbp_index[dt$RBP2]

S[cbind(i, j)] <- dt$Cosine
S[cbind(j, i)] <- dt$Cosine
#diag(S) <- 1

S[is.na(S)] <- 0

D <- 1 - S

pheatmap(
  S,
  clustering_method = "ward.D2",
  color = colorRampPalette(c("navy", "white", "firebrick"))(100),
  show_rownames = TRUE,
  show_colnames = TRUE,
  border_color = NA,
  main = "shRNA Similarity – Both Events"
)



```

### MDS

```{r fig.height=20, fig.width=20}

pair_df <- result

rbp_stats <- rbind(
  pair_df[, .(RBP = RBP1, n_pairs = n_rows)],
  pair_df[, .(RBP = RBP2, n_pairs = n_rows)]
)[
  ,
  .(n_pairs = sum(n_pairs)),
  by = RBP
]

mds <- cmdscale(D, k = 2, eig = TRUE)

mds_df <- data.frame(
  RBP  = rownames(D),
  MDS1 = mds$points[, 1],
  MDS2 = mds$points[, 2],
  stringsAsFactors = FALSE
)

mds_df <- merge(
  mds_df,
  rbp_stats,
  by = "RBP",
  all.x = TRUE
)

mds_df$n_pairs[is.na(mds_df$n_pairs)] <- 0

ggplot(mds_df, aes(MDS1, MDS2)) +
  geom_point(aes(size = n_pairs), alpha = 0.75) +
  geom_text_repel(
    aes(label = RBP),
    size = 3,
    max.overlaps = 30
  ) +
  scale_size_continuous(range = c(2, 10)) +
  theme_minimal(base_size = 14) +
  labs(
    title = "shRNA Similarity – Both Events",
    size = "Shared eCLIP\ncomparisons"
  )
```

### TSne

```{r fig.height=20, fig.width=20}


# -----------------------------------------
# Choose valid perplexity
# -----------------------------------------
n <- nrow(D)
perp <- min(30, floor((n - 1) / 3))

# -----------------------------------------
# Run t-SNE using distance matrix
# -----------------------------------------
tsne <- Rtsne(
  as.dist(D),        # use distance matrix
  is_distance = TRUE,
  perplexity = perp,
  theta = 0.5,       # speed/accuracy tradeoff
  dims = 2,
  verbose = TRUE
)

tsne_df <- data.frame(
  RBP  = rownames(D),
  tSNE1 = tsne$Y[, 1],
  tSNE2 = tsne$Y[, 2],
  stringsAsFactors = FALSE
)

# -----------------------------------------
# Merge with n_pairs (same as before)
# -----------------------------------------
tsne_df <- merge(
  tsne_df,
  rbp_stats,
  by = "RBP",
  all.x = TRUE
)

tsne_df$n_pairs[is.na(tsne_df$n_pairs)] <- 0

# -----------------------------------------
# Plot
# -----------------------------------------
ggplot(tsne_df, aes(tSNE1, tSNE2)) +
  geom_point(aes(size = n_pairs), alpha = 0.75) +
  geom_text_repel(
    aes(label = RBP),
    size = 3,
    max.overlaps = 30
  ) +
  scale_size_continuous(range = c(2, 10)) +
  theme_minimal(base_size = 14) +
  labs(
    title = "t-SNE – shRNA Similarity (Both Events)",
    size = "Shared eCLIP\ncomparisons"
  )
```

### NRows

```{r fig.height=20, fig.width=20}
# result has: RBP1, RBP2, Wasserstein, n_rows

pair_df <- copy(result)

rbps <- sort(unique(c(pair_df$RBP1, pair_df$RBP2)))
n <- length(rbps)

rbp_index <- setNames(seq_len(n), rbps)

Nmat <- matrix(
  0,
  nrow = n,
  ncol = n,
  dimnames = list(rbps, rbps)
)

i <- rbp_index[pair_df$RBP1]
j <- rbp_index[pair_df$RBP2]

Nmat[cbind(i, j)] <- pair_df$n_rows
Nmat[cbind(j, i)] <- pair_df$n_rows   # symmetric

pheatmap(
  Nmat,
  clustering_method = "ward.D2",
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  color = colorRampPalette(c("navy", "white", "firebrick"))(100),
  show_rownames = TRUE,
  show_colnames = TRUE,
  border_color = NA,
  main = "shRNA #Pairs – Both Events"
)

```

# G. Alternative Method (eCLIP)

## Increased

```{r}
library(data.table)

dt <- fread(
  "/mnt/scratch/home/alexandre/Projects/CHARM/intermediary_files/FullSimilarity_oddsratinc.csv"
)

dt <- unique(dt)

# -----------------------------------------------------
# Force Target1 < Target2 ordering
# -----------------------------------------------------
dt[, `:=`(
  Target_low  = pmin(Target1, Target2),
  Target_high = pmax(Target1, Target2)
)]

dt[, RBP_low  := ifelse(Target1 < Target2, RBP1, RBP2)]
dt[, RBP_high := ifelse(Target1 < Target2, RBP2, RBP1)]

# -----------------------------------------------------
# Aggregate per Target pair
# -----------------------------------------------------
result <- dt[
  ,
  .(
    Cosine = mean(Cosine, na.rm = TRUE),
    n_rows = uniqueN(c(RBP_low, RBP_high))
  ),
  by = .(Target1 = Target_low, Target2 = Target_high)
]

```

### Heatmap

```{r fig.height=20, fig.width=20}
dt <- copy(result)

targets <- sort(unique(c(dt$Target1, dt$Target2)))
n <- length(targets)

target_index <- setNames(seq_len(n), targets)

S <- matrix(
  NA_real_,
  n, n,
  dimnames = list(targets, targets)
)

i <- target_index[dt$Target1]
j <- target_index[dt$Target2]

S[cbind(i, j)] <- dt$Cosine
S[cbind(j, i)] <- dt$Cosine

S[is.na(S)] <- 0
D <- 1 - S

pheatmap(
  S,
  clustering_method = "ward.D2",
  color = colorRampPalette(c("navy", "white", "firebrick"))(100),
  show_rownames = TRUE,
  show_colnames = TRUE,
  border_color = NA,
  main = "Target Similarity – Increased Events"
)

```

### MDS

```{r fig.height=20, fig.width=20}
pair_df <- copy(result)

target_stats <- rbind(
  pair_df[, .(Target = Target1, n_pairs = n_rows)],
  pair_df[, .(Target = Target2, n_pairs = n_rows)]
)[
  ,
  .(n_pairs = sum(n_pairs)),
  by = Target
]

mds <- cmdscale(D, k = 2, eig = TRUE)

mds_df <- data.frame(
  Target = rownames(D),
  MDS1 = mds$points[, 1],
  MDS2 = mds$points[, 2],
  stringsAsFactors = FALSE
)

mds_df <- merge(
  mds_df,
  target_stats,
  by = "Target",
  all.x = TRUE
)

mds_df$n_pairs[is.na(mds_df$n_pairs)] <- 0

ggplot(mds_df, aes(MDS1, MDS2)) +
  geom_point(aes(size = n_pairs), alpha = 0.75) +
  geom_text_repel(
    aes(label = Target),
    size = 3,
    max.overlaps = 30
  ) +
  scale_size_continuous(range = c(2, 10)) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Target Similarity – Increased Events",
    size = "Shared RBP\ncomparisons"
  )

```

### TSne

```{r fig.height=20, fig.width=20}


# -----------------------------------------
# Choose valid perplexity
# -----------------------------------------
n <- nrow(D)
perp <- min(30, floor((n - 1) / 3))

# -----------------------------------------
# Run t-SNE using distance matrix
# -----------------------------------------
tsne <- Rtsne(
  as.dist(D),        # use distance matrix
  is_distance = TRUE,
  perplexity = perp,
  theta = 0.5,       # speed/accuracy tradeoff
  dims = 2,
  verbose = TRUE
)

tsne_df <- data.frame(
  RBP  = rownames(D),
  tSNE1 = tsne$Y[, 1],
  tSNE2 = tsne$Y[, 2],
  stringsAsFactors = FALSE
)

# -----------------------------------------
# Merge with n_pairs (same as before)
# -----------------------------------------
tsne_df <- merge(
  tsne_df,
  rbp_stats,
  by = "RBP",
  all.x = TRUE
)

tsne_df$n_pairs[is.na(tsne_df$n_pairs)] <- 0

# -----------------------------------------
# Plot
# -----------------------------------------
ggplot(tsne_df, aes(tSNE1, tSNE2)) +
  geom_point(aes(size = n_pairs), alpha = 0.75) +
  geom_text_repel(
    aes(label = RBP),
    size = 3,
    max.overlaps = 30
  ) +
  scale_size_continuous(range = c(2, 10)) +
  theme_minimal(base_size = 14) +
  labs(
    title = "t-SNE – Target Similarity (Increased Events)",
    size = "Shared eCLIP\ncomparisons"
  )
```

### NRows

```{r fig.height=20, fig.width=20}
pair_df <- copy(result)

targets <- sort(unique(c(pair_df$Target1, pair_df$Target2)))
n <- length(targets)

target_index <- setNames(seq_len(n), targets)

Nmat <- matrix(
  0,
  n, n,
  dimnames = list(targets, targets)
)

i <- target_index[pair_df$Target1]
j <- target_index[pair_df$Target2]

Nmat[cbind(i, j)] <- pair_df$n_rows
Nmat[cbind(j, i)] <- pair_df$n_rows

pheatmap(
  (Nmat),
  clustering_method = "ward.D2",
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  color = colorRampPalette(c("navy", "white", "firebrick"))(100),
  show_rownames = TRUE,
  show_colnames = TRUE,
  border_color = NA,
  main = "eCLIP #Pairs – Increased Events"
)

```


## Decreased

```{r}
library(data.table)

dt <- fread(
  "/mnt/scratch/home/alexandre/Projects/CHARM/intermediary_files/FullSimilarity_oddsratdec.csv"
)

dt <- unique(dt)

# -----------------------------------------------------
# Force Target1 < Target2 ordering
# -----------------------------------------------------
dt[, `:=`(
  Target_low  = pmin(Target1, Target2),
  Target_high = pmax(Target1, Target2)
)]

dt[, RBP_low  := ifelse(Target1 < Target2, RBP1, RBP2)]
dt[, RBP_high := ifelse(Target1 < Target2, RBP2, RBP1)]

# -----------------------------------------------------
# Aggregate per Target pair
# -----------------------------------------------------
result <- dt[
  ,
  .(
    Cosine = mean(Cosine, na.rm = TRUE),
    n_rows = uniqueN(c(RBP_low, RBP_high))
  ),
  by = .(Target1 = Target_low, Target2 = Target_high)
]
```

### Heatmap

```{r fig.height=20, fig.width=20}
dt <- copy(result)

targets <- sort(unique(c(dt$Target1, dt$Target2)))
n <- length(targets)

target_index <- setNames(seq_len(n), targets)

S <- matrix(
  NA_real_,
  n, n,
  dimnames = list(targets, targets)
)

i <- target_index[dt$Target1]
j <- target_index[dt$Target2]

S[cbind(i, j)] <- dt$Cosine
S[cbind(j, i)] <- dt$Cosine

S[is.na(S)] <- 0
D <- 1 - S

pheatmap(
  S,
  clustering_method = "ward.D2",
  color = colorRampPalette(c("navy", "white", "firebrick"))(100),
  show_rownames = TRUE,
  show_colnames = TRUE,
  border_color = NA,
  main = "Target Similarity – Decreased Events"
)

```

### MDS

```{r fig.height=20, fig.width=20}
pair_df <- copy(result)

target_stats <- rbind(
  pair_df[, .(Target = Target1, n_pairs = n_rows)],
  pair_df[, .(Target = Target2, n_pairs = n_rows)]
)[
  ,
  .(n_pairs = sum(n_pairs)),
  by = Target
]

mds <- cmdscale(D, k = 2, eig = TRUE)

mds_df <- data.frame(
  Target = rownames(D),
  MDS1 = mds$points[, 1],
  MDS2 = mds$points[, 2],
  stringsAsFactors = FALSE
)

mds_df <- merge(
  mds_df,
  target_stats,
  by = "Target",
  all.x = TRUE
)

mds_df$n_pairs[is.na(mds_df$n_pairs)] <- 0

ggplot(mds_df, aes(MDS1, MDS2)) +
  geom_point(aes(size = n_pairs), alpha = 0.75) +
  geom_text_repel(
    aes(label = Target),
    size = 3,
    max.overlaps = 30
  ) +
  scale_size_continuous(range = c(2, 10)) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Target Similarity – Decreased Events",
    size = "Shared RBP\ncomparisons"
  )

```


### TSne

```{r fig.height=20, fig.width=20}


# -----------------------------------------
# Choose valid perplexity
# -----------------------------------------
n <- nrow(D)
perp <- min(30, floor((n - 1) / 3))

# -----------------------------------------
# Run t-SNE using distance matrix
# -----------------------------------------
tsne <- Rtsne(
  as.dist(D),        # use distance matrix
  is_distance = TRUE,
  perplexity = perp,
  theta = 0.5,       # speed/accuracy tradeoff
  dims = 2,
  verbose = TRUE
)

tsne_df <- data.frame(
  RBP  = rownames(D),
  tSNE1 = tsne$Y[, 1],
  tSNE2 = tsne$Y[, 2],
  stringsAsFactors = FALSE
)

# -----------------------------------------
# Merge with n_pairs (same as before)
# -----------------------------------------
tsne_df <- merge(
  tsne_df,
  rbp_stats,
  by = "RBP",
  all.x = TRUE
)

tsne_df$n_pairs[is.na(tsne_df$n_pairs)] <- 0

# -----------------------------------------
# Plot
# -----------------------------------------
ggplot(tsne_df, aes(tSNE1, tSNE2)) +
  geom_point(aes(size = n_pairs), alpha = 0.75) +
  geom_text_repel(
    aes(label = RBP),
    size = 3,
    max.overlaps = 30
  ) +
  scale_size_continuous(range = c(2, 10)) +
  theme_minimal(base_size = 14) +
  labs(
    title = "t-SNE – Target Similarity (Decreased Events)",
    size = "Shared eCLIP\ncomparisons"
  )
```

### NRows

```{r fig.height=20, fig.width=20}
pair_df <- copy(result)

targets <- sort(unique(c(pair_df$Target1, pair_df$Target2)))
n <- length(targets)

target_index <- setNames(seq_len(n), targets)

Nmat <- matrix(
  0,
  n, n,
  dimnames = list(targets, targets)
)

i <- target_index[pair_df$Target1]
j <- target_index[pair_df$Target2]

Nmat[cbind(i, j)] <- pair_df$n_rows
Nmat[cbind(j, i)] <- pair_df$n_rows

pheatmap(
  (Nmat),
  clustering_method = "ward.D2",
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  color = colorRampPalette(c("navy", "white", "firebrick"))(100),
  show_rownames = TRUE,
  show_colnames = TRUE,
  border_color = NA,
  main = "eCLIP #Pairs – Decreased Events"
)

```

## Both

```{r}
library(data.table)

dt1 <- fread(
  "/mnt/scratch/home/alexandre/Projects/CHARM/intermediary_files/FullSimilarity_oddsratinc.csv"
)

dt2 <- fread(
  "/mnt/scratch/home/alexandre/Projects/CHARM/intermediary_files/FullSimilarity_oddsratdec.csv"
)

dt <- rbind(dt1, dt2)
#dt <- unique(dt)
# -----------------------------------------------------
# Force Target1 < Target2 ordering
# -----------------------------------------------------
dt[, `:=`(
  Target_low  = pmin(Target1, Target2),
  Target_high = pmax(Target1, Target2)
)]

dt[, RBP_low  := ifelse(Target1 < Target2, RBP1, RBP2)]
dt[, RBP_high := ifelse(Target1 < Target2, RBP2, RBP1)]

# -----------------------------------------------------
# Aggregate per Target pair
# -----------------------------------------------------
result <- dt[
  ,
  .(
    Cosine = mean(Cosine, na.rm = TRUE),
    n_rows = uniqueN(c(RBP_low, RBP_high))
  ),
  by = .(Target1 = Target_low, Target2 = Target_high)
]

```

### Heatmap

```{r fig.height=20, fig.width=20}
dt <- copy(result)

targets <- sort(unique(c(dt$Target1, dt$Target2)))
n <- length(targets)

target_index <- setNames(seq_len(n), targets)

S <- matrix(
  NA_real_,
  n, n,
  dimnames = list(targets, targets)
)

i <- target_index[dt$Target1]
j <- target_index[dt$Target2]

S[cbind(i, j)] <- dt$Cosine
S[cbind(j, i)] <- dt$Cosine

S[is.na(S)] <- 0
D <- 1 - S

pheatmap(
  S,
  clustering_method = "ward.D2",
  color = colorRampPalette(c("navy", "white", "firebrick"))(100),
  show_rownames = TRUE,
  show_colnames = TRUE,
  border_color = NA,
  main = "Target Similarity – Both Events"
)

```

### MDS

```{r fig.height=20, fig.width=20}
pair_df <- copy(result)

target_stats <- rbind(
  pair_df[, .(Target = Target1, n_pairs = n_rows)],
  pair_df[, .(Target = Target2, n_pairs = n_rows)]
)[
  ,
  .(n_pairs = sum(n_pairs)),
  by = Target
]

mds <- cmdscale(D, k = 2, eig = TRUE)

mds_df <- data.frame(
  Target = rownames(D),
  MDS1 = mds$points[, 1],
  MDS2 = mds$points[, 2],
  stringsAsFactors = FALSE
)

mds_df <- merge(
  mds_df,
  target_stats,
  by = "Target",
  all.x = TRUE
)

mds_df$n_pairs[is.na(mds_df$n_pairs)] <- 0

ggplot(mds_df, aes(MDS1, MDS2)) +
  geom_point(aes(size = n_pairs), alpha = 0.75) +
  geom_text_repel(
    aes(label = Target),
    size = 3,
    max.overlaps = 30
  ) +
  scale_size_continuous(range = c(2, 10)) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Target Similarity – Both Events",
    size = "Shared RBP\ncomparisons"
  )

```


### TSne

```{r fig.height=20, fig.width=20}


# -----------------------------------------
# Choose valid perplexity
# -----------------------------------------
n <- nrow(D)
perp <- min(30, floor((n - 1) / 3))

# -----------------------------------------
# Run t-SNE using distance matrix
# -----------------------------------------
tsne <- Rtsne(
  as.dist(D),        # use distance matrix
  is_distance = TRUE,
  perplexity = perp,
  theta = 0.5,       # speed/accuracy tradeoff
  dims = 2,
  verbose = TRUE
)

tsne_df <- data.frame(
  RBP  = rownames(D),
  tSNE1 = tsne$Y[, 1],
  tSNE2 = tsne$Y[, 2],
  stringsAsFactors = FALSE
)

# -----------------------------------------
# Merge with n_pairs (same as before)
# -----------------------------------------
tsne_df <- merge(
  tsne_df,
  rbp_stats,
  by = "RBP",
  all.x = TRUE
)

tsne_df$n_pairs[is.na(tsne_df$n_pairs)] <- 0

# -----------------------------------------
# Plot
# -----------------------------------------
ggplot(tsne_df, aes(tSNE1, tSNE2)) +
  geom_point(aes(size = n_pairs), alpha = 0.75) +
  geom_text_repel(
    aes(label = RBP),
    size = 3,
    max.overlaps = 30
  ) +
  scale_size_continuous(range = c(2, 10)) +
  theme_minimal(base_size = 14) +
  labs(
    title = "t-SNE – Target Similarity (Both Events)",
    size = "Shared eCLIP\ncomparisons"
  )
```

### NRows

```{r fig.height=20, fig.width=20}
pair_df <- copy(result)

targets <- sort(unique(c(pair_df$Target1, pair_df$Target2)))
n <- length(targets)

target_index <- setNames(seq_len(n), targets)

Nmat <- matrix(
  0,
  n, n,
  dimnames = list(targets, targets)
)

i <- target_index[pair_df$Target1]
j <- target_index[pair_df$Target2]

Nmat[cbind(i, j)] <- pair_df$n_rows
Nmat[cbind(j, i)] <- pair_df$n_rows

pheatmap(
  (Nmat),
  clustering_method = "ward.D2",
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  color = colorRampPalette(c("navy", "white", "firebrick"))(100),
  show_rownames = TRUE,
  show_colnames = TRUE,
  border_color = NA,
  main = "eCLIP #Pairs – Both Events"
)

```


# H. Intron Retention

## File Preparation Part 1 

First, I will prepare the files by maintaining only the psi that maximise odds ratio.

```{r fig.height=20, fig.width=20}
Charm.object.binding.IR <- readRDS("../data/Binding_IR_HEPG2.RDS")

filter_binding_dpsi <- function(binding_obj) {

  keep_pattern <- "(maximised oddsratinc|maximised oddsratdec)"

  out <- binding_obj

  for (rbp in names(out)) {
    for (tgt in names(out[[rbp]])) {

      dpsi_names <- names(out[[rbp]][[tgt]])
      if (is.null(dpsi_names)) next

      keep <- dpsi_names[
        grepl(keep_pattern, dpsi_names, ignore.case = TRUE) |
        grepl("(^|[^0-9])0\\.05([^0-9]|$)", dpsi_names) |
        grepl("(^|[^0-9])0\\.10([^0-9]|$)", dpsi_names)
      ]

      if (length(keep) == 0) {
        out[[rbp]][[tgt]] <- NULL
      } else {
        out[[rbp]][[tgt]] <- out[[rbp]][[tgt]][keep]
      }
    }

    if (length(out[[rbp]]) == 0) {
      out[[rbp]] <- NULL
    }
  }

  out
}
filter_binding_keep_oddsrat <- function(binding_obj) {

  out <- binding_obj

  for (rbp in names(out)) {
    for (tgt in names(out[[rbp]])) {
      for (dpsi in names(out[[rbp]][[tgt]])) {

        dpsi_block <- out[[rbp]][[tgt]][[dpsi]]

        # If oddsrat does not exist, drop this dPSI
        if (!is.list(dpsi_block) || !"oddsrat" %in% names(dpsi_block)) {
          out[[rbp]][[tgt]][[dpsi]] <- NULL
          next
        }

        # Safely extract raw event counts (may be NULL)
        inc_events <- NULL
        dec_events <- NULL

        if ("raw" %in% names(dpsi_block)) {
          raw_block <- dpsi_block[["raw"]]

          if (is.list(raw_block)) {
            inc_events <- raw_block[["IncreasedTargetEvents"]]
            dec_events <- raw_block[["DecreasedTargetEvents"]]
          }
        }

        # Replace dPSI block with trimmed content
        out[[rbp]][[tgt]][[dpsi]] <- list(
          oddsrat = dpsi_block[["oddsrat"]],
          IncreasedTargetEvents = inc_events,
          DecreasedTargetEvents = dec_events
        )
      }

      # Drop empty targets
      if (length(out[[rbp]][[tgt]]) == 0) {
        out[[rbp]][[tgt]] <- NULL
      }
    }

    # Drop empty RBPs
    if (length(out[[rbp]]) == 0) {
      out[[rbp]] <- NULL
    }
  }

  out
}

Charm.binding.IR.filtered <- filter_binding_dpsi(Charm.object.binding.IR)
Charm.binding.IR.filtered <- filter_binding_keep_oddsrat(Charm.binding.IR.filtered)
#rm(Charm.object.binding.ES)
```

Great, now I will calculate how much signal each pair of RBP and Target has, at selected dPSIs.

```{r}
summarise_oddsrat_abs_sum <- function(binding_obj) {

  out <- list()

  for (rbp in names(binding_obj)) {
    out[[rbp]] <- list()

    for (tgt in names(binding_obj[[rbp]])) {
      out[[rbp]][[tgt]] <- list()

      for (dpsi in names(binding_obj[[rbp]][[tgt]])) {

        dpsi_block <- binding_obj[[rbp]][[tgt]][[dpsi]]
        if (!is.list(dpsi_block)) next

        oddsrat_block <- dpsi_block[["oddsrat"]]
        if (is.null(oddsrat_block)) next

        # --- extract event counts safely ---
        inc_tbl <- dpsi_block[["IncreasedTargetEvents"]]
        dec_tbl <- dpsi_block[["DecreasedTargetEvents"]]

        n_inc <- if (is.data.frame(inc_tbl) && "Value" %in% colnames(inc_tbl))
          inc_tbl$Value[1] else 0

        n_dec <- if (is.data.frame(dec_tbl) && "Value" %in% colnames(dec_tbl))
          dec_tbl$Value[1] else 0

        dpsi_out <- list()

        # -------------------------
        # Decreased
        # -------------------------
        df_dec <- oddsrat_block[["oddsratdec"]]
        if (is.data.frame(df_dec) && n_dec > 0) {
          dpsi_out[["oddsratdec"]] <-
            sum(abs(df_dec$Value), na.rm = TRUE) * (n_dec^(1/2))
        }

        # -------------------------
        # Increased
        # -------------------------
        df_inc <- oddsrat_block[["oddsratinc"]]
        if (is.data.frame(df_inc) && n_inc > 0) {
          dpsi_out[["oddsratinc"]] <-
            sum(abs(df_inc$Value), na.rm = TRUE) * (n_inc^(1/2))
        }

        if (length(dpsi_out) > 0) {
          out[[rbp]][[tgt]][[dpsi]] <- dpsi_out
        }
      }

      if (length(out[[rbp]][[tgt]]) == 0)
        out[[rbp]][[tgt]] <- NULL
    }

    if (length(out[[rbp]]) == 0)
      out[[rbp]] <- NULL
  }

  out
}


Charm.binding.oddsrat.sum <- summarise_oddsrat_abs_sum(Charm.binding.IR.filtered)

extract_oddsrat_values <- function(oddsrat_obj) {

  inc_vals <- numeric()
  dec_vals <- numeric()

  for (rbp in names(oddsrat_obj)) {
    for (tgt in names(oddsrat_obj[[rbp]])) {
      for (dpsi in names(oddsrat_obj[[rbp]][[tgt]])) {

        block <- oddsrat_obj[[rbp]][[tgt]][[dpsi]]

        if (!is.null(block$oddsratinc)) {
          inc_vals <- c(inc_vals, block$oddsratinc)
        }
        if (!is.null(block$oddsratdec)) {
          dec_vals <- c(dec_vals, block$oddsratdec)
        }
      }
    }
  }

  list(
    oddsratinc = inc_vals,
    oddsratdec = dec_vals
  )
}

plot_oddsrat_densities <- function(oddsrat_obj,
                                   inc_color = "red",
                                   dec_color = "blue",
                                   bw = "nrd0") {

  vals <- extract_oddsrat_values(oddsrat_obj)

  if (length(vals$oddsratinc) < 2 || length(vals$oddsratdec) < 2) {
    stop("Not enough values to compute densities.")
  }

  dens_inc <- density(log10(vals$oddsratinc), bw = bw)
  dens_dec <- density(log10(vals$oddsratdec), bw = bw)

  # ---- Increased ----
  plot(
    dens_inc,
    col = inc_color,
    lwd = 3,
    main = "Distribution of |Odds Ratio| – Increased Events",
    xlab = "log10|Odds Ratio| Integral",
    ylab = "Density"
  )+abline(v=4.5, col="black")

  # ---- Decreased ----
  plot(
    dens_dec,
    col = dec_color,
    lwd = 3,
    main = "Distribution of |Odds Ratio| – Decreased Events",
    xlab = "log10|Odds Ratio| Integral",
    ylab = "Density"
  )
}

plot_oddsrat_densities(Charm.binding.oddsrat.sum)


```

 Counting the before and after:
 
```{r}
filter_and_select_top_oddsrat <- function(oddsrat_obj,
                                          log10_threshold = 6) {

  filtered <- list()
  n_inc <- 0L
  n_dec <- 0L

  for (rbp in names(oddsrat_obj)) {
    for (tgt in names(oddsrat_obj[[rbp]])) {

      best_inc <- NULL
      best_inc_dpsi <- NULL

      best_dec <- NULL
      best_dec_dpsi <- NULL

      for (dpsi in names(oddsrat_obj[[rbp]][[tgt]])) {

        block <- oddsrat_obj[[rbp]][[tgt]][[dpsi]]

        ## oddsratinc
        if (!is.null(block$oddsratinc) &&
            is.finite(block$oddsratinc) &&
            log10(block$oddsratinc) > log10_threshold) {

          if (is.null(best_inc) || block$oddsratinc > best_inc) {
            best_inc <- block$oddsratinc
            best_inc_dpsi <- dpsi
          }
        }

        ## oddsratdec
        if (!is.null(block$oddsratdec) &&
            is.finite(block$oddsratdec) &&
            log10(block$oddsratdec) > log10_threshold) {

          if (is.null(best_dec) || block$oddsratdec > best_dec) {
            best_dec <- block$oddsratdec
            best_dec_dpsi <- dpsi
          }
        }
      }

      ## Store results (at most 1 inc + 1 dec)
      if (!is.null(best_inc) || !is.null(best_dec)) {

        filtered[[rbp]][[tgt]] <- list()

        if (!is.null(best_inc)) {
          filtered[[rbp]][[tgt]][[best_inc_dpsi]]$oddsratinc <- best_inc
          n_inc <- n_inc + 1L
        }

        if (!is.null(best_dec)) {
          filtered[[rbp]][[tgt]][[best_dec_dpsi]]$oddsratdec <- best_dec
          n_dec <- n_dec + 1L
        }
      }
    }
  }

  list(
    counts = c(
      oddsratinc = n_inc,
      oddsratdec = n_dec
    ),
    filtered_object = filtered
  )
}

res <- filter_and_select_top_oddsrat(
  Charm.binding.oddsrat.sum,
  log10_threshold = -20
)

res$counts
#res$filtered_object



```

```{r}
filtered_object <- res[["filtered_object"]]

all_targets <- unlist(
  lapply(filtered_object, names),
  use.names = FALSE
)

unique_targets <- (unique(all_targets))

length(unique(all_targets))

filtered_object <- res$filtered_object

filtered_object <- lapply(
  filtered_object,
  function(rbp_block) {
    rbp_block[names(rbp_block) %in% unique_targets]
  }
)


```

```{r}
filtered_object <- res[["filtered_object"]]
flatten_oddsrat <- function(filtered_object,
                            direction = c("oddsratinc", "oddsratdec")) {

  direction <- match.arg(direction)
  out <- list()
  k <- 1L

  for (rbp in names(filtered_object)) {
    for (tgt in names(filtered_object[[rbp]])) {
      for (dpsi in names(filtered_object[[rbp]][[tgt]])) {

        vec <- filtered_object[[rbp]][[tgt]][[dpsi]][[direction]]

        if (is.null(vec) || length(vec) == 0) next
        if (!is.numeric(vec)) next

        out[[k]] <- data.table(
          RBP    = rbp,
          Target = tgt,
          dPSI   = dpsi,
          Value  = vec
        )
        k <- k + 1L
      }
    }
  }

  rbindlist(out, use.names = TRUE)
}

oddsratinc_df <- flatten_oddsrat(
  filtered_object,
  direction = "oddsratinc"
)

oddsratdec_df <- flatten_oddsrat(
  filtered_object,
  direction = "oddsratdec"
)

```


```{r fig.height=20, fig.width=20}
library(data.table)
library(pheatmap)

setnames(
  oddsratinc_df,
  old = "Value",
  new = "oddsratinc"
)

setnames(
  oddsratdec_df,
  old = "Value",
  new = "oddsratdec"
)

rbp_target_heatmap_raw <- function(
  df,
  value_col,
  agg_fun = c("mean", "median"),
  clustering_method = "ward.D2",
  clustering_distance = "euclidean",
  show_rownames = TRUE,
  show_colnames = TRUE,
  main = "RBP × Target heatmap (raw values)"
) {

  agg_fun <- match.arg(agg_fun)
  dt <- as.data.table(df)

  stopifnot(all(c("RBP", "Target", value_col) %in% colnames(dt)))

  # -------------------------------------------------
  # Aggregate ONLY if needed (no transformation)
  # -------------------------------------------------
  agg <- dt[
    ,
    .(
      value = switch(
        agg_fun,
        mean   = mean(get(value_col), na.rm = TRUE),
        median = median(get(value_col), na.rm = TRUE)
      )
    ),
    by = .(RBP, Target)
  ]

  # -------------------------------------------------
  # Build matrix
  # -------------------------------------------------
  rbps    <- sort(unique(agg$RBP))
  targets <- sort(unique(agg$Target))

  S <- matrix(
    NA_real_,
    nrow = length(rbps),
    ncol = length(targets),
    dimnames = list(rbps, targets)
  )

  i <- match(agg$RBP, rbps)
  j <- match(agg$Target, targets)

  S[cbind(i, j)] <- agg$value

  # -------------------------------------------------
  # Plot
  # -------------------------------------------------
  pheatmap(
    log10(S),
    clustering_method = clustering_method,
    clustering_distance_rows = clustering_distance,
    clustering_distance_cols = clustering_distance,
    color = colorRampPalette(c("navy", "white", "firebrick"))(100),
    show_rownames = show_rownames,
    show_colnames = show_colnames,
    border_color = NA,
    main = main
  )

  # S = your matrix/dataframe (rows = RBPs, cols = targets)

max_values <- apply(S, 2, max, na.rm = TRUE)       # maximum per column
max_rows   <- apply(S, 2, function(col) {
  rownames(S)[which.max(col)]
})

# Combine into a dataframe
max_per_column <- data.frame(
  Target = colnames(S),
  MaxValue = max_values,
  MaxRBP = max_rows
)

print(max_per_column)

  
  invisible(S)
}

#oddsratinc_df_alter <- 

rbp_target_heatmap_raw(
  oddsratinc_df,
  value_col = "oddsratinc",
  main = "RBP × Target – Oddsratinc (raw)"
)

rbp_target_heatmap_raw(
  oddsratdec_df,
  value_col = "oddsratdec",
  main = "RBP × Target – Oddsratdec (raw)"
)

```

```{r fig.height=20, fig.width=20}
library(data.table)
library(pheatmap)

rbp_target_heatmap_raw <- function(
  df,
  value_col,
  agg_fun = c("mean", "median"),
  clustering_method = "ward.D2",
  clustering_distance = "euclidean",
  show_rownames = TRUE,
  show_colnames = TRUE,
  main = "RBP × Target heatmap (raw values)"
) {

  agg_fun <- match.arg(agg_fun)
  dt <- as.data.table(df)

  stopifnot(all(c("RBP", "Target", value_col) %in% colnames(dt)))

  # -------------------------------------------------
  # Aggregate ONLY if needed (no transformation)
  # -------------------------------------------------
  agg <- dt[
    ,
    .(
      value = switch(
        agg_fun,
        mean   = mean(get(value_col), na.rm = TRUE),
        median = median(get(value_col), na.rm = TRUE)
      )
    ),
    by = .(RBP, Target)
  ]

  # -------------------------------------------------
  # Build matrix
  # -------------------------------------------------
  rbps    <- sort(unique(agg$RBP))
  targets <- sort(unique(agg$Target))

  S <- matrix(
    NA_real_,
    nrow = length(rbps),
    ncol = length(targets),
    dimnames = list(rbps, targets)
  )

  i <- match(agg$RBP, rbps)
  j <- match(agg$Target, targets)

  S[cbind(i, j)] <- agg$value

  # -------------------------------------------------
  # Plot
  # -------------------------------------------------
S_log <- log10(S + 1)  # add 1 to avoid log(0)
S_z <- scale(S_log, center = TRUE, scale = TRUE)

# Keep as a matrix for pheatmap
S_z <- as.matrix(S_z)

  
pheatmap(
  S_z,
  cluster_rows = TRUE,   # cluster RBPs
  cluster_cols = TRUE,  # keep targets in the original order
  clustering_method = "ward.D2",
  color = colorRampPalette(c("navy", "white", "firebrick"))(100),
  show_rownames = TRUE,
  show_colnames = TRUE,
  border_color = NA,
  main = main
)

  # S = your matrix/dataframe (rows = RBPs, cols = targets)

max_values <- apply(S, 2, max, na.rm = TRUE)       # maximum per column
max_rows   <- apply(S, 2, function(col) {
  rownames(S)[which.max(col)]
})

# Combine into a dataframe
max_per_column <- data.frame(
  Target = colnames(S),
  MaxValue = max_values,
  MaxRBP = max_rows
)

print(max_per_column)

  
  invisible(S)
}

#oddsratinc_df_alter <- 

rbp_target_heatmap_raw(
  oddsratinc_df,
  value_col = "oddsratinc",
  main = "RBP × Target – log10Oddsratinc (Target zScore)"
)

rbp_target_heatmap_raw(
  oddsratdec_df,
  value_col = "oddsratdec",
  main = "RBP × Target – log10Oddsratdec (Target zScore)"
)

```

```{r fig.height=20, fig.width=20}
rbp_target_pca <- function(
  df,
  value_col,
  agg_fun = c("mean", "median"),
  log_transform = TRUE,
  zscore_columns = TRUE,
  min_non_na = 2,
  main = "PCA – RBPs (Targets as genes)"
) {

  agg_fun <- match.arg(agg_fun)
  dt <- as.data.table(df)

  stopifnot(all(c("RBP", "Target", value_col) %in% colnames(dt)))

  # -------------------------------------------------
  # Aggregate values per RBP–Target
  # -------------------------------------------------
  agg <- dt[
    ,
    .(
      value = switch(
        agg_fun,
        mean   = mean(get(value_col), na.rm = TRUE),
        median = median(get(value_col), na.rm = TRUE)
      )
    ),
    by = .(RBP, Target)
  ]

  # -------------------------------------------------
  # Build matrix: rows = RBPs, cols = Targets
  # -------------------------------------------------
  S <- dcast(agg, RBP ~ Target, value.var = "value")

  rbp_names <- S$RBP
  S$RBP <- NULL
  S <- as.matrix(S)
  rownames(S) <- rbp_names

  # Replace NAs with 0 (or choose another strategy if desired)
  S[is.na(S)] <- 0

  # -------------------------------------------------
  # Transformations
  # -------------------------------------------------
  if (log_transform) {
    S <- log10(S + 1)
  }

  if (zscore_columns) {
    S <- scale(S, center = TRUE, scale = TRUE)
  }

  S <- as.matrix(S)

  # -------------------------------------------------
  # Remove zero-variance Targets
  # -------------------------------------------------
  keep_cols <- apply(S, 2, function(x)
    sum(is.finite(x)) >= min_non_na && sd(x, na.rm = TRUE) > 0
  )

  S <- S[, keep_cols, drop = FALSE]

  # -------------------------------------------------
  # PCA (RBPs are observations)
  # -------------------------------------------------
  pca_res <- prcomp(S, center = FALSE, scale. = FALSE)

  var_explained <- (pca_res$sdev^2) / sum(pca_res$sdev^2)

  pca_df <- data.frame(
    RBP = rownames(S),
    PC1 = pca_res$x[, 1],
    PC2 = pca_res$x[, 2]
  )

  # -------------------------------------------------
  # Plot
  # -------------------------------------------------
  p <- ggplot(pca_df, aes(PC1, PC2)) +
    geom_point(size = 4, alpha = 0.8) +
    geom_text_repel(aes(label = RBP), size = 3) +
    theme_minimal(base_size = 14) +
    labs(
      title = main,
      x = paste0("PC1 (", round(var_explained[1] * 100, 1), "%)"),
      y = paste0("PC2 (", round(var_explained[2] * 100, 1), "%)")
    )

  print(p)

  invisible(list(
    matrix = S,
    pca = pca_res,
    variance = var_explained,
    plot = p
  ))
}

rbp_target_pca(
  oddsratinc_df,
  value_col = "oddsratinc",
  main = "PCA – RBPs (Oddsratinc)"
)

rbp_target_pca(
  oddsratdec_df,
  value_col = "oddsratdec",
  main = "PCA – RBPs (Oddsratdec)"
)

```


## Filtering
 
```{r}
filter_original_binding_by_oddsrat <- function(original_binding,
                                               filtered_object) {

  out <- list()

  for (rbp in names(filtered_object)) {
    if (!rbp %in% names(original_binding)) next

    out[[rbp]] <- list()

    for (tgt in names(filtered_object[[rbp]])) {
      if (!tgt %in% names(original_binding[[rbp]])) next

      out[[rbp]][[tgt]] <- list()

      for (dpsi in names(filtered_object[[rbp]][[tgt]])) {
        if (!dpsi %in% names(original_binding[[rbp]][[tgt]])) next

        oddsrat_block <- original_binding[[rbp]][[tgt]][[dpsi]][["oddsrat"]]
        if (is.null(oddsrat_block)) next

        keep_dirs <- filtered_object[[rbp]][[tgt]][[dpsi]]
        new_oddsrat <- list()

        for (dir in names(keep_dirs)) {
          if (dir %in% names(oddsrat_block)) {
            new_oddsrat[[dir]] <- oddsrat_block[[dir]]
          }
        }

        if (length(new_oddsrat) > 0) {
          out[[rbp]][[tgt]][[dpsi]] <- list(
            oddsrat = new_oddsrat
          )
        }
      }

      if (length(out[[rbp]][[tgt]]) == 0) {
        out[[rbp]][[tgt]] <- NULL
      }
    }

    if (length(out[[rbp]]) == 0) {
      out[[rbp]] <- NULL
    }
  }

  out
}

Charm.object.binding.filtered <- filter_original_binding_by_oddsrat(
  original_binding = Charm.object.binding.IR,
  filtered_object  = res$filtered_object
)


filter_original_binding_by_oddsrat <- function(original_binding,
                                               filtered_object) {

  out <- list()

  for (rbp in names(filtered_object)) {
    if (!rbp %in% names(original_binding)) next

    out[[rbp]] <- list()

    for (tgt in names(filtered_object[[rbp]])) {
      if (!tgt %in% names(original_binding[[rbp]])) next

      out[[rbp]][[tgt]] <- list()

      for (dpsi in names(filtered_object[[rbp]][[tgt]])) {
        if (!dpsi %in% names(original_binding[[rbp]][[tgt]])) next

        block <- original_binding[[rbp]][[tgt]][[dpsi]]

        oddsrat_block <- block[["oddsrat"]]
        raw_block     <- block[["raw"]]

        if (is.null(oddsrat_block) || is.null(raw_block)) next

        # -------------------------------------------------
        # Extract event scalars correctly
        # -------------------------------------------------
        inc_events <- raw_block[["IncreasedTargetEvents"]][["Value"]]
        dec_events <- raw_block[["DecreasedTargetEvents"]][["Value"]]

        keep_dirs <- filtered_object[[rbp]][[tgt]][[dpsi]]
        new_oddsrat <- list()

        for (dir in names(keep_dirs)) {

          if (!dir %in% names(oddsrat_block)) next

          df <- oddsrat_block[[dir]]
          if (!is.data.frame(df) || !"Value" %in% colnames(df)) next

          df_new <- df

          # -------------------------------------------------
          # Multiply correctly
          # -------------------------------------------------
          if (dir == "oddsratinc" && !is.null(inc_events)) {
            df_new$Value <- df_new$Value * inc_events
          }

          if (dir == "oddsratdec" && !is.null(dec_events)) {
            df_new$Value <- df_new$Value * dec_events
          }

          new_oddsrat[[dir]] <- df_new
        }

        if (length(new_oddsrat) > 0) {
          out[[rbp]][[tgt]][[dpsi]] <- list(
            oddsrat = new_oddsrat
          )
        }
      }

      if (length(out[[rbp]][[tgt]]) == 0) {
        out[[rbp]][[tgt]] <- NULL
      }
    }

    if (length(out[[rbp]]) == 0) {
      out[[rbp]] <- NULL
    }
  }

  out
}

Charm.object.binding.filtered <- filter_original_binding_by_oddsrat(
  original_binding = Charm.object.binding.IR,
  filtered_object  = res$filtered_object
)


saveRDS(Charm.object.binding.filtered, "../intermediary_files/Charm.binding.filtered.both_new_full_IR_HEPG2.Rds")
```
 
 
## Creating python format

```{r}
oddsrat_list_to_dataframe <- function(filtered_obj,
                                      n_pos = 500,
                                      prefix = "pos_") {

  rows <- vector("list", 50000)
  idx <- 1L

  for (rbp in names(filtered_obj)) {
    for (tgt in names(filtered_obj[[rbp]])) {
      for (dpsi in names(filtered_obj[[rbp]][[tgt]])) {

        block <- filtered_obj[[rbp]][[tgt]][[dpsi]]

        ## Safety
        if (is.null(block$oddsrat))
          next

        ## ---------- oddsratinc ----------
        if (!is.null(block$oddsrat$oddsratinc)) {

          v <- block$oddsrat$oddsratinc$Value

          if (length(v) != n_pos)
            stop(paste("Unexpected length for oddsratinc:",
                       rbp, tgt, dpsi))

          rows[[idx]] <- c(
            RBP = rbp,
            Target = tgt,
            dPSI = dpsi,
            Direction = "oddsratinc",
            setNames(as.list(v), paste0(prefix, seq_len(n_pos)))
          )

          idx <- idx + 1L
        }

        ## ---------- oddsratdec ----------
        if (!is.null(block$oddsrat$oddsratdec)) {

          v <- block$oddsrat$oddsratdec$Value

          if (length(v) != n_pos)
            stop(paste("Unexpected length for oddsratdec:",
                       rbp, tgt, dpsi))

          rows[[idx]] <- c(
            RBP = rbp,
            Target = tgt,
            dPSI = dpsi,
            Direction = "oddsratdec",
            setNames(as.list(v), paste0(prefix, seq_len(n_pos)))
          )

          idx <- idx + 1L
        }
      }
    }
  }

  if (idx == 1L)
    stop("No oddsrat entries found — check structure again")

  df <- as.data.frame(
    do.call(rbind, rows[seq_len(idx - 1)]),
    stringsAsFactors = FALSE
  )

  ## Convert positional columns back to numeric
  pos_cols <- grep(paste0("^", prefix), colnames(df))
  df[, pos_cols] <- lapply(df[, pos_cols], as.numeric)

  df
}
oddsrat_df <- oddsrat_list_to_dataframe(
  Charm.object.binding.filtered,
  n_pos = 500   # or 500 for IR
)

data.table::fwrite(
  oddsrat_df,
  "../intermediary_files/Charm.binding.filtered.both_new_full_IR_HEPG2.csv",
  sep = "\t"
)

```


